---
output: github_document
---

```{r}
#| label: set-up
#| echo: false

knitr::opts_chunk$set(
  eval = FALSE, echo = TRUE
)
```

[![Zenodo](https://img.shields.io/badge/Code%20DOI-10.5281/zenodo.14903161-blue)](https://doi.org/10.5281/zenodo.14903161) [![medRxiv](https://img.shields.io/badge/medRxiv-10.1101/2025.04.10.25325611-blue)](https://doi.org/10.1101/2025.04.10.25325611)

## Using model-based evaluation to interpret variation in infectious disease forecast performance

Katharine Sherratt (1), Rok Grah (2), Bastian Prasse (2), Friederike Becker (3), Jamie McLean (1), Sam Abbott (1), Sebastian Funk (1)

(1) Centre for Mathematical Modelling of Infectious Diseases, London School of Hygiene & Tropical Medicine
(2) European Centre for Disease Prevention and Control
(3) Institute of Statistics, Karlsruhe Institute of Technology

<details>
<summary>Abstract</summary>

```{r abstract, child="report/quarto/_abstract.qmd", eval=TRUE}
```

</details>

--------------------------------------------------------------------------------

### Getting started

#### Read

Read the work as it stands:

- The full paper (background, methods, results, discussion) is in [report/manuscript.qmd](./report/manuscript.qmd), which assembles the section files in [report/quarto/](./report/quarto/).
- The supplement is in [report/quarto/supplement/\_supplement.qmd](./report/quarto/supplement/_supplement.qmd).
- The former pre-print is on [medRxiv](https://doi.org/10.1101/2025.04.10.25325611); note, this is out of date.

#### Reproduce

To re-run the analysis end to end, without editing anything:

1. Restore the package environment ([renv](https://rstudio.github.io/renv/articles/renv.html)):

   ```{r}
      renv::restore()
   ```

2. Data is already in `data/`.
   To re-download it from public sources, see the [data/README](data/README.md).

3. Score forecasts on the log and natural scales (writes to `data/`):

   ```{r}
      source(here("R", "process-score.R"))
   ```

4. Assemble scores with explanatory variables:

   ```{r}
      source(here("R", "process-data.R"))
   ```

5. Fit the GAMM to the weighted interval score (writes to `output/`):

   ```{r}
      source(here("R", "analysis-model.R"))
   ```

6. Render the manuscript (includes the results section and supplement):

   ```{r}
      quarto::quarto_render("report/manuscript.qmd")
   ```

   `analysis-model.R` must be run before rendering --- it is not sourced by the results section.

#### Explore

A guide to the codebase:

- `R/` --- analysis and utility scripts:
  - `process-score.R` scores forecasts; `process-data.R` joins scores to explanatory variables; `analysis-model.R` fits the GAMM; `analysis-descriptive.R` builds summary tables; `plot-model-results.R` and `plot-model-flow.R` make the figures; `dag-check.R` defines the confounding DAG.
  - `utils-data.R`, `utils-metadata.R`, `utils-variants.R` --- data access, model metadata, and variant-phase classification.
  - `R/sensitivity/` --- robustness checks (autocorrelation, link function, log-response model, model-building notebook).
- `data/` --- forecasts (`covid19-forecast-hub-europe.parquet`), observed incidence, populations, model classification, and computed scores.
- `report/quarto/` --- the manuscript section files; `report/quarto/supplement/` the supplement.
- `output/` --- fitted model results and diagnostic plots, under scale-named subdirectories (`log/`, `log-resp/`, `natural/`).

--------------------------------------------------------------------------------

### Project docs

- [Authorship](https://docs.google.com/spreadsheets/d/18mt6c47MCzLdMAKth6Bv2PC7b10KYgWieExHHWLnP3Q/edit?gid=0#gid=0)
- [Slide deck](https://docs.google.com/presentation/d/1BSdTEuZ_zKdU8tBFuRMmP7GwHht1D0oZSkaFWovz9ao/edit?slide=id.p#slide=id.p) --- high-level context
