---
title: "Standardising forecast scores by epidemic regime"
format:
  html:
    toc: true
    df-print: kable
execute:
  echo: false
  message: false
  warning: false
  cache: true
---

```{r}
#| label: setup

library(here)
library(tidyverse)
library(mgcv)
library(marginaleffects)

theme_set(theme_classic())

source(here("R", "process-data.R"))

scores <- process_data(scoring_scale = "log") |>
  filter(!grepl("EuroCOVIDhub-ensemble", Model)) |>
  filter(Horizon == 1)

# Weekly growth, computed on the distinct observation series rather than on
# forecast rows, where consecutive rows are the same week and a different model.
weekly <- scores |>
  distinct(Location, epi_target, target_end_date, Incidence, pop) |>
  arrange(Location, epi_target, target_end_date) |>
  group_by(Location, epi_target) |>
  mutate(Growth = log((Incidence + 0.5) / (lag(Incidence) + 0.5))) |>
  ungroup() |>
  select(Location, epi_target, target_end_date, Growth)

regime_breaks <- c(-Inf, -0.2, -0.05, 0.05, 0.2, Inf)
regime_labels <- c("Falling fast", "Falling", "Stable", "Rising", "Rising fast")

d <- scores |>
  left_join(weekly, by = c("Location", "epi_target", "target_end_date")) |>
  mutate(
    IncidencePK = log((Incidence + 0.5) / pop * 1e5),
    Regime = cut(Growth, regime_breaks, labels = regime_labels),
    Epi_target = factor(epi_target),
    Model = factor(Model),
    Location = factor(Location)
  ) |>
  filter(is.finite(wis), is.finite(IncidencePK), is.finite(Growth))
```

## The problem

Forecast evaluations report a mean score per target type, and compare those means.
Cases and deaths are scored separately, and deaths usually score better.

That mean is taken over whatever mix of epidemic conditions the evaluation period happened to contain.
If score depends on epidemic state, and the state distribution differs between the things being compared, the comparison is confounded by the mix rather than by anything about forecasting.

This is analogous to a traditional epidemiologist's crude rate comparison:

| Mortality | Forecast evaluation |
|---|---|
| Crude death rate | Mean score over all targets |
| Age stratum | Epidemic regime (level, growth) |
| Age-specific rate | Regime-specific mean score |
| Reference population | Reference regime distribution |
| Age-standardised rate | Regime-standardised mean score |

When stratum-specific rates are proportional across groups, a single standardised summary is appropriate.
However, when they cross or diverge, the summary hides informative variation (i.e. effect modification), and the strata should be shown instead.

## Data and stratification

Weekly COVID-19 case and death forecasts submitted to the European Forecast Hub, scored with the weighted interval score on the log scale (LWIS), horizon one-week only.
The Hub ensemble is excluded.

Epidemic regime is the weekly log change in observed incidence, cut at ±0.05 and ±0.2, computed within country and outcome.
Growth is dimensionless, so its strata mean the same thing for cases and deaths.
Incidence level is not comparable across outcomes and is not used for stratification (but see below).

```{r}
#| label: tbl-regime-mix
#| tbl-cap: "Distribution of target weeks across epidemic regimes, percentage total, by outcome"

regime_mix <- d |>
  distinct(Location, epi_target, target_end_date, Regime) |>
  count(epi_target, Regime) |>
  group_by(epi_target) |>
  mutate(pct = round(100 * n / sum(n), 1)) |>
  ungroup() |>
  select(-n) |>
  pivot_wider(names_from = epi_target, values_from = pct)

regime_mix
```

If columns in @tbl-regime-mix show very different distributions, this suggests standardisation is a useful adjustment before comparison.

## Crude, stratum-specific, and standardised distributions

```{r}
#| label: tbl-crude
#| tbl-cap: "Crude comparison: mean LWIS per forecast, by outcome."

crude <- d |>
  group_by(epi_target) |>
  summarise(crude = mean(wis), `N forecasts` = n(), .groups = "drop")

crude |> mutate(`Crude mean LWIS` = round(crude, 4))
```

```{r}
#| label: tbl-strata
#| tbl-cap: "Regime-specific mean LWIS by outcome; ratios below one favour deaths."

strata <- d |>
  group_by(epi_target, Regime) |>
  summarise(mean_wis = mean(wis), forecasts = n(), .groups = "drop")

strata_wide <- strata |>
  select(-forecasts) |>
  pivot_wider(names_from = epi_target, values_from = mean_wis) |>
  mutate(ratio_deaths_cases = Deaths / Cases) |>
  mutate(across(where(is.numeric), \(x) round(x, 4)))

strata_wide
```

For standardisation, any fixed reference works; we reference the pooled regime distribution across both outcomes, so neither outcome is privileged.

```{r}
#| label: tbl-standardised
#| tbl-cap: "Directly standardised mean LWIS, weighting regime-specific means by the pooled regime distribution across both outcomes."

reference <- d |>
  distinct(Location, epi_target, target_end_date, Regime) |>
  count(Regime) |>
  mutate(weight = n / sum(n)) |>
  select(Regime, weight)

standardised <- strata |>
  left_join(reference, by = "Regime") |>
  group_by(epi_target) |>
  summarise(standardised = sum(weight * mean_wis), .groups = "drop") |>
  left_join(crude, by = "epi_target") |>
  select(epi_target, crude, standardised) |>
  mutate(across(where(is.numeric), \(x) round(x, 4)))

standardised
```

### Interpretation

Compare @tbl-crude with @tbl-standardised to see how much of the crude gap is regime mix.
Then read the ratio column of @tbl-strata to see whether a single standardised number is defensible.

A roughly constant ratio across regimes means the outcomes differ by a common factor, and either the crude or the standardised comparison will adequately summarise it.
A ratio that changes across regimes means the difference between outcomes depends on epidemic state, and no single number represents it.
In that case the standardised value would be substantively misleading (in the same way an age-standardised rate ratio is when the age-specific ratios cross).

Here the ratio is close to constant and the two regime distributions are close to identical, so standardisation over phase is defensible and leaves the comparison where it started.
That is a useful negative: it rules phase out as an explanation for the case-death gap, which the crude comparison alone cannot do.

## Model-based standardisation

Direct standardisation on a single stratifying variable is difficult to support with sparse stratum-specific data, and cannot cross-stratify for an incidence level, country, and variant constant at the same time.
The model-based equivalent is g-computation: fit score as a function of the target-generating variables, predict every target under a common regime, and average.

```{r}
#| label: fit-target-model

m_target <- bam(
  wis ~
    Epi_target +
    te(IncidencePK, Growth, by = Epi_target, k = c(5, 5)) +
    s(VariantPhase, bs = "re") +
    s(Location, bs = "re"),
  data = d,
  family = tw(link = "log"),
  discrete = TRUE,
  nthreads = 4
)
```

```{r}
#| label: tbl-gcomp
#| tbl-cap: "Model-based standardisation: mean predicted LWIS with growth set to each regime midpoint for every target, holding all other covariates at their observed values."

regime_points <- c(-0.35, -0.12, 0, 0.12, 0.35)

gcomp <- map_dfr(regime_points, \(g) {
  avg_predictions(
    m_target,
    newdata = transform(d, Growth = g),
    by = "Epi_target",
    type = "response"
  ) |>
    as_tibble() |>
    mutate(Growth = g)
}) |>
  select(Epi_target, Growth, estimate, conf.low, conf.high) |>
  mutate(across(where(is.numeric), \(x) round(x, 4)))

gcomp
```

This holds incidence level, country, and variant at their observed values while moving only growth, so it separates epidemic phase from the level and place it usually travels with.

## Where standardisation runs out

Growth is dimensionless and means the same thing for both outcomes, which is what made the sections above possible.
Incidence level does not: five deaths per 100,000 in a week is a catastrophe and five cases per 100,000 is nothing.
No reference distribution over level means the same thing on both sides of the comparison, so the level component of difficulty cannot be standardised across outcomes at all.

This is not a technicality, because level is where the difficulty gradient sits.
Moving phase across its full range in @tbl-gcomp changes the score by a fifth; moving level changes it by an order of magnitude.

Within a single outcome, level does decompose.
Rate carries epidemic severity and count carries observation noise, and the two separate because population spans two orders of magnitude across the 32 countries.

```{r}
#| label: tbl-rate-count
#| tbl-cap: "Mean LWIS for death forecasts by weekly rate per 100,000 and by weekly count. Reading across a row holds rate roughly fixed and varies count; reading down a column holds count and varies rate. Target counts are distinct country-weeks."

rate_count <- d |>
  filter(epi_target == "Deaths") |>
  mutate(
    count = cut(Incidence, c(-1, 5, 20, 100, Inf)),
    rate  = cut(Incidence / pop * 1e5, c(-1, 0.5, 1, 5, Inf))
  ) |>
  group_by(rate, count) |>
  summarise(
    mean_wis = round(mean(wis), 4),
    targets = n_distinct(Location, target_end_date),
    .groups = "drop"
  ) |>
  pivot_wider(names_from = count, values_from = c(mean_wis, targets))

rate_count
```

Both gradients are present in @tbl-rate-count, and they carry different meanings.

The row gradient is observation noise.
A country recording three deaths a week bounces between one and six for reasons unconnected to the epidemic, and that variation is large in the space the score is computed in.
Small counts are therefore harder, not easier, which is the opposite of what a compression argument would predict.

The column gradient is a mix of epidemic difficulty and the scoring transform.
`process-score.R` normalises to per 100,000 and then applies `log(x + 1)`, so the offset sits on a rate rather than a count.
Where the rate is well below one per 100,000, `log(r + 1)` is close to `r` and almost no transformation happens; where it is in the hundreds, the transform behaves as intended.
Death scores are therefore closer to an absolute error measure and case scores closer to a relative one, which is enough on its own to produce part of the gap.

Two cells are too thin to read: the highest rate band at counts below five, and at counts of 5 to 20.
The second is structurally empty rather than merely sparse, since a rate above five per 100,000 at a count under twenty requires a population below 400,000.

## Conclusion

Standardisation transfers cleanly to forecast evaluation when the stratifying variable has a common scale across the things being compared.
Epidemic phase qualifies, and applying it here rules phase out as an explanation for the case-death difference.

Epidemic level does not qualify, and level is where the difficulty actually lives.
The apparent superiority of death forecasts is therefore not a claim standardisation can settle, and neither is its converse.

The useful output is the diagnostic rather than an adjusted number: a comparison that looks standardisable, an argument for why one component of it is and the other is not, and a scoring-scale artefact surfaced by asking the question.

## Limitations

Exploratory. No pre-specified hypothesis, and the regime cutpoints were chosen for interpretability rather than derived.

LWIS mixes calibration and sharpness, so a higher score means a forecast was scored worse, not that the target was harder in any independent sense.

Growth is measured on observed data as reported, so revisions and reporting artefacts enter the stratification.
Deaths lag cases, so a death week classified as stable may follow a volatile case week; the regimes are contemporaneous with the target, not with the epidemic.

The g-computation in @tbl-gcomp assumes positivity, and it does not hold everywhere.
Setting growth to 0.35 for every target includes country-weeks whose incidence level never co-occurs with fast growth, so some predictions come from unsupported regions of the tensor surface.
Restricting to rows inside each outcome's observed joint range, or reporting how many fall outside it, would bound this.

Model composition is not adjusted for.
If the set of models forecasting in one regime differs from another, regime-specific means absorb that difference.
Adding `s(Model, bs = "re")` to `m_target` tests it, at the cost of no longer being a purely target-side description.

Horizon one only.
If regime effects work through turning points, they should grow with horizon, which is a check rather than an assumption.
