```{r}
#| label: results-set-up
#| include: false

library(here)
library(dplyr)
library(readr)
library(tidyr)
library(tibble)
library(purrr)
library(ggplot2)
library(patchwork)
library(kableExtra)
theme_set(theme_classic())
knitr::opts_chunk$set(
  eval = TRUE, echo = FALSE,
  message = FALSE, warning = FALSE
)
source(here("R", "process-data.R"))
source(here("R", "analysis-descriptive.R"))
source(here("R", "plot-model-results.R"))
## to re-run and save model results:
# source(here("R", "analysis-model.R")) 
# model_wis(scoring_scale = "log", output_dir = "output")

scores <- process_data(scoring_scale = "log")
ensemble <- scores |>
  filter(grepl("EuroCOVIDhub-ensemble", Model))
scores <- scores |>
  filter(!grepl("EuroCOVIDhub-ensemble", Model))
```

```{r}
#| label: results-participation-text

max_forecasts <- 104 * 2 * 4 * 32 # 104 weeks * 2 epi outcomes * 4 horizons * 32 countries
n_models <- length(unique(scores$Model))
n_forecasts <- nrow(scores)
n_teams <- distinct(scores, Model) |>
  separate(Model, into = c("team", "model"), sep = "-") |>
  count(team)
teams_models <- summary(n_teams$n)

# Coverage of the target matrix: how many of the available targets drew any
# forecast, and how many models
cells_covered <- scores |>
  distinct(epi_target, Location, forecast_date, Horizon) |>
  nrow()
p_cells <- round(cells_covered / max_forecasts * 100)
models_per_cell <- scores |>
  count(epi_target, Location, forecast_date, Horizon) |>
  pull(n) |>
  summary()
```

We evaluated `r n_models` forecasting models from `r nrow(n_teams)` modelling teams participating in the European COVID-19 Forecast Hub over 2021-2023.

#### Participant characteristics

```{r}
#| label: tbl-models
#| tbl-cap: >-
#|   Characteristics of models sampled from the European COVID-19 Forecast Hub, March 2021-2023.
#|   Models (%) shows the number of models and the percentage of that column's total, grouped by which epidemiological outcome(s) a model forecast: cases only, deaths only, both, and all included models combined.
#|   Participation is the median (IQR) across models in each group of the percentage of the 26,624 available forecast targets a model submitted for.

# Overall participation
model_forecasts <- scores |>
  group_by(Model) |>
  summarise(n_forecasts = n(),
  p_max = n_forecasts / max_forecasts * 100)
model_participation <- select(model_forecasts, Model, Participation = p_max)
model_forecasts <- round(summary(model_forecasts$p_max), 0)
# Model structures
structures <- scores |>
  select(Model, Method, agreement) |>
  distinct()
structure_count <- structures |>
  group_by(Method) |>
  summarise(agreement = sum(agreement),
            n = n(),
            p = agreement / n)
# Named lookups so the text below tracks the data rather than hardcoding counts
n_by_method <- setNames(structure_count$n, as.character(structure_count$Method))
n_disagree <- sum(!structures$agreement)
p_disagree <- round(n_disagree / nrow(structures) * 100)
p_by_method <- scores |>
  count(Method) |>
  mutate(p = round(n / sum(n) * 100)) |>
  (\(x) setNames(x$p, as.character(x$Method)))()
# Semi-mechanistic models that drew at least one statistical vote
semi_mech_votes <- classify_models(return_majority = FALSE) |>
  filter(model %in% structures$Model[structures$Method == "Semi-mechanistic"])
n_semi_mech_split <- n_distinct(
  semi_mech_votes$model[semi_mech_votes$classification == "Statistical"]
)

# Epi targets
model_epi_target <- scores |>
  group_by(Method, Model, .drop = TRUE) |>
  distinct(epi_target, .keep_all = TRUE) |>
  summarise(model_epi_target = paste(sort(unique(epi_target)), collapse = " & ")) |>
  mutate(`Model structure` = as.character(Method)) |>
  left_join(model_participation, by = "Model")
gtsummary::tbl_summary(
  model_epi_target,
  include = c(`Model structure`, Participation),
  by = model_epi_target,
  statistic = list(Participation ~ "{median} ({p25}, {p75})"),
  digits = list(Participation ~ 1),
  label = list(Participation ~ "Participation (%)")
) |>
  gtsummary::add_overall(last = TRUE, col_label = "All included, N = {N}")
```

Forecasters could participate in any of 26,624 available targets (104 weeks, 32 countries, 4 horizons, 2 epidemiological outcomes).
Among them, forecasters covered `r p_cells`% of available targets, with a median of `r models_per_cell[["Median"]]` models forecasting each covered target (IQR `r models_per_cell[["1st Qu."]]`-`r models_per_cell[["3rd Qu."]]`, maximum `r models_per_cell[["Max."]]`).
Coverage was uneven across countries, weeks and model structures (Supplement).
Median participation was `r model_forecasts[["Median"]]`% of available targets (IQR `r model_forecasts[["1st Qu."]]`-`r model_forecasts[["3rd Qu."]]`%, range `r model_forecasts[["Min."]]` to `r model_forecasts[["Max."]]`).

We classified `r n_by_method[["Mechanistic"]]` models as mechanistic, `r n_by_method[["Statistical"]]` as statistical, `r n_by_method[["Semi-mechanistic"]]` as semi-mechanistic, `r n_by_method[["Agent-based"]]` as agent-based and `r n_by_method[["Judgement"]]` as human judgement (@tbl-models; Supplement).
Raters disagreed on `r n_disagree` (`r p_disagree`%) classifications, most often for semi-mechanistic models, `r n_semi_mech_split` of which drew one vote for a statistical structure.
Statistical models contributed the largest share of predictions (`r p_by_method[["Statistical"]]`%), partly because the Hub baseline model submitted for every country in almost every week.
Mechanistic and semi-mechanistic models contributed `r p_by_method[["Mechanistic"]]`% and `r p_by_method[["Semi-mechanistic"]]`%, and agent-based and judgement models under 2% each.

#### Forecast performance

```{r}
#| label: fig-error-vs-obs
#| fig-cap: >-
#|   Log-transform weighted interval score (LWIS), against the magnitude of observed incidence per 100,000 population, for COVID-19 cases and deaths across 32 European countries.
#|   Lower LWIS indicates better performance.
#|   Ribbons show the 5th, 25th, and 75th percentiles of LWIS (y axis), summarised within increasing bands of observed incidence spaced across the log scale (x axis).
#|   Ribbons are coloured by forecast horizon.
#|   Both axes are shown on the log scale.
#| fig-height: 4
#| fig-width: 8

plot_error_vs_obs_bands(scores)
```

```{r}
#| label: results-describe-performance

# Median LWIS by outcome and by variant phase, to describe unadjusted
# performance without fitting anything
median_by_target <- scores |>
  group_by(epi_target) |>
  summarise(median = sprintf("%.2f", median(wis, na.rm = TRUE))) |>
  (\(x) setNames(x$median, x$epi_target))()
median_by_phase <- scores |>
  group_by(VariantPhase) |>
  summarise(median = median(wis, na.rm = TRUE)) |>
  arrange(median)
phase_best <- as.character(median_by_phase$VariantPhase[1])
phase_worst <- as.character(median_by_phase$VariantPhase[nrow(median_by_phase)])
median_by_horizon <- scores |>
  group_by(epi_target, Horizon) |>
  summarise(median = sprintf("%.2f", median(wis, na.rm = TRUE)), .groups = "drop") |>
  (\(x) setNames(x$median, paste(x$epi_target, x$Horizon)))()

# The Hub ensemble is excluded from the fit because it is built from the
# forecasts it would be compared against; it is kept here as a reference point
ensemble_median <- ensemble |>
  group_by(epi_target) |>
  summarise(median = sprintf("%.2f", median(wis, na.rm = TRUE))) |>
  (\(x) setNames(x$median, x$epi_target))()
beat_ensemble <- scores |>
  group_by(epi_target, Model) |>
  summarise(median = median(wis, na.rm = TRUE), .groups = "drop") |>
  left_join(
    ensemble |>
      group_by(epi_target) |>
      summarise(ensemble = median(wis, na.rm = TRUE)),
    by = "epi_target"
  ) |>
  group_by(epi_target) |>
  summarise(n_models = n(), n_better = sum(median < ensemble))
```

Aggregated performance varied more with the epidemiological process being predicted than between forecasters.
Case forecasts scored a median LWIS of `r median_by_target[["Cases"]]` against `r median_by_target[["Deaths"]]` for deaths.
For deaths in particular, scores varied non-linearly with the magnitude of the outcome being predicted (@fig-error-vs-obs).
Across variant phases, forecasts were most accurate during the `r phase_best` phase and least accurate during `r phase_worst`, the period of steepest change in reported incidence.
Scores worsened with forecast horizon (@fig-error-vs-obs), from a median LWIS of `r median_by_horizon[["Cases 1"]]` at one week to `r median_by_horizon[["Cases 4"]]` at four weeks for cases, and `r median_by_horizon[["Deaths 1"]]` to `r median_by_horizon[["Deaths 4"]]` for deaths.
Against the Hub ensemble, constructed from the median of the forecasts we assess in this analysis, `r beat_ensemble$n_better[beat_ensemble$epi_target == "Cases"]` of `r beat_ensemble$n_models[beat_ensemble$epi_target == "Cases"]` models achieved a lower median LWIS for cases (ensemble `r ensemble_median[["Cases"]]`) and `r beat_ensemble$n_better[beat_ensemble$epi_target == "Deaths"]` of `r beat_ensemble$n_models[beat_ensemble$epi_target == "Deaths"]` for deaths (ensemble `r ensemble_median[["Deaths"]]`).

```{r}
#| label: results-sparsity

# How thin the sample becomes once cross-classified by the categorical
# covariates, i.e. what a stratified analysis would have to work with
strata_vars <- c("epi_target", "Horizon", "Location", "Trend",
                 "VariantPhase", "CountryTargets", "Method")
strata_possible <- prod(sapply(scores[strata_vars], n_distinct))
strata_obs <- count(scores, across(all_of(strata_vars)))
strata_filled <- round(nrow(strata_obs) / strata_possible * 100)
strata_median <- median(strata_obs$n)
strata_thin <- round(mean(strata_obs$n < 5) * 100)
```

#### Effect of model structure

Participation between model structures was unevenly distributed across forecast targets.
Handling this with a stratification approach would produce sparse data, given both the range of targets, and number and type of covariates we wish to adjust for.
To demonstrate, cross-classifying forecasts by the categorical covariates alone (epidemiological outcome, location, epidemic trend, variant phase, number of countries targeted) gives `r format(strata_possible, big.mark = ",")` possible strata, of which `r strata_filled`% contain any forecast.
Among these, each non-empty stratum holds a median of only `r strata_median` forecasts and `r strata_thin`% hold fewer than five.
We report model-based effect estimates as a ratio against the grand-mean LWIS for each factor, under a sum-to-zero constraint; a ratio of 0.9 represents an LWIS 10% lower (better) than the group average.
We report raw partial effects in the Supplement.

```{r}
#| label: results-scores-results

results <- readRDS(here("output", "log", "results.rds"))
table_effects <- results$effects |> 
  mutate(upper_97.5_text = if_else(
           upper_97.5 < 0,
           paste0("(", round(upper_97.5, 2), ")"),
           as.character(round(upper_97.5, 2))
         )) |>
  mutate(value_ci = paste0(round(value, 2),
                           " (", round(lower_2.5, 2), "-",
                           upper_97.5_text, ")"),
         # exponentiated: multiplicative ratio vs grand-mean LWIS (1 = average).
         # Fixed to 2 dp so ratios at or near 1 print as 1.00, not 1
         ratio_ci = sprintf("%.2f (%.2f-%.2f)",
                            exp(value), exp(lower_2.5), exp(upper_97.5)),
         group = paste(model, group, sep = "_")) |>
  column_to_rownames("group")

effects_comp <- results$effects |> 
  pivot_longer(cols = c(value, lower_2.5, upper_97.5, se)) |> 
  pivot_wider(names_from = model) |> 
  mutate(diff = Adjusted - Unadjusted,
  diff_percent = diff / Unadjusted * 100) |> 
  filter(group_var %in% c("Method"))
```

After adjustment, no structural approach dominated (@tbl-structure, @fig-structure-effects).
The largest shifts under adjustment were among agent-based and human judgement models, while adjustment shrank the standard errors throughout.

```{r}
#| label: tbl-structure
#| tbl-cap: >-
#|   Adjusting for features of the forecasting context alters the interpretation of variation.
#|   Estimated effects of model structure on COVID-19 forecast performance, where a ratio below 1 indicates better-than-average performance.
#|   Performance measured as LWIS (weighted interval score of log-transformed forecasts and observations), and estimated from univariate (unadjusted) and adjusted generalised additive mixed models.
#|   Adjustment included: the interaction of model structure with epidemiological outcome; weekly incidence, trend, variant phase, location and forecast horizon; whether a model forecast for one or many locations; and an individul model-specific effect.
#|   95% CI = 95% confidence interval.

print_table2(results$effects)
```

```{r}
#| label: results-method-target

mbt <- results$method_by_target
mbt_ratio <- setNames(
  sprintf(
    "%.2f (%.2f-%.2f)",
    exp(mbt$value), exp(mbt$lower_2.5), exp(mbt$upper_97.5)
  ),
  paste(mbt$Method, mbt$Epi_target, sep = "_")
)
# case minus death effect on the log scale; negative indicates a structure
# performed relatively better on cases than on deaths
mbt_diff <- vapply(
  split(mbt, mbt$Method),
  \(x) x$value[x$Epi_target == "Cases"] - x$value[x$Epi_target == "Deaths"],
  numeric(1)
)
mbt_widest <- names(which.max(abs(mbt_diff)))
# How much data the widest-separating structure rests on, per outcome
mbt_widest_n <- scores |>
  filter(Method == mbt_widest) |>
  group_by(epi_target) |>
  summarise(
    n_models = n_distinct(Model),
    n_forecasts = format(n(), big.mark = ","),
    .groups = "drop"
  )
```

Differences between model structures were modified by epidemiological outcome (@fig-structure-effects; Supplement).
The widest separation was among `r tolower(mbt_widest)` models, which were the worst-performing model structure for case forecasts but the best for death forecasts (cases ratio `r mbt_ratio[[paste0(mbt_widest, "_Cases")]]`; deaths ratio `r mbt_ratio[[paste0(mbt_widest, "_Deaths")]]`), while their pooled estimate sits close to the grand mean.
This contrast rests on `r mbt_widest_n$n_models[mbt_widest_n$epi_target == "Cases"]` models and `r mbt_widest_n$n_forecasts[mbt_widest_n$epi_target == "Cases"]` forecasts for cases, and `r mbt_widest_n$n_models[mbt_widest_n$epi_target == "Deaths"]` models and `r mbt_widest_n$n_forecasts[mbt_widest_n$epi_target == "Deaths"]` forecasts for deaths, so the structure-level estimate is close to a statement about individual models, even with a model-specific effect in the fit.

Other model structures performed more similarly between epidemiological outcomes, performing marginally better on cases than on deaths.
We noted that differences between outcomes largely offset when averaged.
An averaged estimate therefore gives close to zero for most structures, whether or not outcome-specific differences exist.

```{r}
#| label: fig-structure-effects
#| fig-cap: >-
#|   No model structure performed meaningfully differently from the average after adjustment for features of the epidemiological prediction target.
#|   Showing performance ratio (95% CI) by model structure, as the exponentiated partial effect relative to the grand-mean LWIS; a ratio below 1 indicates better-than-average performance.
#|   LWIS = weighted interval score of log-transformed forecasts and observations.
#| fig-height: 3
#| fig-width: 9

plot_method_target(results$method_by_target)
```

#### Wider influences on forecast performance

Epidemiological processes generating the forecast target explained more variation in performance than features of the forecasting process.
Targets were most predictable during stable periods of the epidemic curve (ratio `r table_effects["Adjusted_Stable","ratio_ci"]`) and least during increasing trends (ratio `r table_effects["Adjusted_Increasing","ratio_ci"]`).
Predictability was better in the earlier Alpha and Delta phases (ratios `r table_effects["Adjusted_Alpha","ratio_ci"]` and `r table_effects["Adjusted_Delta","ratio_ci"]`) and worst under Omicron BA.1 (ratio `r table_effects["Adjusted_Omicron-BA.1","ratio_ci"]`), and death forecasts scored lower than case forecasts (ratio `r table_effects["Adjusted_Deaths","ratio_ci"]`).
We report the total set of modelled effects in the Supplement.

Most of the difference between individual model performance was left unexplained (@fig-model-variation A).
We accounted for whether models forecast for one or multiple countries, with 19 models forecasting for a single country and 29 for multiple countries (of which only 2 kept the same number of countries throughout, see Supplement).
Both estimates were shrunk to the average (single-country ratio `r table_effects["Adjusted_Single-country","ratio_ci"]`; multi-country ratio `r table_effects["Adjusted_Multi-country","ratio_ci"]`), so geographic scope explained none of the difference between models.
Forecast horizon entered the model as a smooth fitted separately for each model, so it has no single pooled effect; the crude gradient across horizons is given above.

```{r}
#| label: results-ranks
#| include: false

# TODO fix panel B
ranks <- rank_models(results$effects)
rank_summary <- summarise_ranks(ranks)
```

We compared this remaining variation between individual models in terms of ranked model order.
Ranking models by their performance before and after adjustment gives a Spearman correlation of `r round(rank_summary$spearman, 2)`.
`r rank_summary$n_moved` of `r rank_summary$n` models moved at least `r rank_summary$threshold` places, with the furthest moving `r rank_summary$max_change` places (@fig-model-variation B).

```{r}
#| label: fig-model-variation
#| fig-cap: >-
#|   Individual models differ in performance after adjustment, so which models look best depends on how the evaluation is designed.
#|   (A) Adjusted performance ratio (95% CI) for each model, ordered by effect and labelled by the model's unadjusted rank.
#|   The ratio is the exponentiated partial effect relative to the grand-mean LWIS, where a ratio below 1 indicates better-than-average performance, adjusted for model structure, geographic scope, forecast horizon, epidemic trend, variant phase and location.
#|   (B) Rank of each model before and after adjustment, with the diagonal marking no change in rank, coloured by model structure.
#|   LWIS = weighted interval score of log-transformed forecasts and observations.
#| fig-height: 8
#| fig-width: 11

plot_model_variation(results$effects, ranks = ranks)
```

#### Sensitivity analysis

Repeating the analysis on the natural scale of scores, and under alternative covariate sets, changed no estimate materially (Supplement).
The exception was the choice of error family: refitting with a Gaussian in place of the Tweedie changed the size of every structure-by-outcome estimate and the direction of two of the five.
Specifically, this widened the case-death difference for human judgement models, removed it for agent-based models, and reversed it for semi-mechanistic models.
We explain this where using a Gaussian family is dominated by the largest scores, which fall disproportionately among case forecasts.
