```{r}
#| label: results-set-up
#| include: false

library(here)
library(dplyr)
library(readr)
library(tidyr)
library(tibble)
library(purrr)
library(ggplot2)
library(patchwork)
library(kableExtra)
theme_set(theme_classic())
knitr::opts_chunk$set(
  eval = TRUE, echo = FALSE,
  message = FALSE, warning = FALSE
)
source(here("R", "process-data.R"))
source(here("R", "analysis-descriptive.R"))
source(here("R", "plot-model-results.R"))
## to re-run and save model results:
# source(here("R", "analysis-model.R")) 
# model_wis(scoring_scale = "log", output_dir = "output")

scores <- process_data(scoring_scale = "log")
ensemble <- scores |>
  filter(grepl("EuroCOVIDhub-ensemble", Model))
scores <- scores |>
  filter(!grepl("EuroCOVIDhub-ensemble", Model))

# natural-scale scores for descriptive error-vs-observation figure
scores_natural <- process_data(scoring_scale = "natural") |>
  filter(!grepl("EuroCOVIDhub-ensemble", Model))
```

<!---Summary--->
```{r}
#| label: results-participation-text

max_forecasts <- 104 * 2 * 4 * 32 # 104 weeks * 2 epi outcomes * 4 horizons * 32 countries
n_models <- length(unique(scores$Model))
n_forecasts <- nrow(scores)
n_teams <- distinct(scores, Model) |> 
  separate(Model, into = c("team", "model"), sep = "-") |> 
  count(team)
teams_models <- summary(n_teams$n)
```

We evaluated `r n_models` forecasting models from `r nrow(n_teams)` modelling teams participating in the European COVID-19 Forecast Hub.
Models differed both in which targets they forecast and in how well they performed.
We used a generalised additive mixed model to associate performance with model structure, adjusting for the number of countries a model targeted and for the epidemiological context of each target.

#### Forecaster characteristics

```{r}
#| label: results-describe-structures

# Overall participation
model_forecasts <- scores |>
  group_by(Model) |>
  summarise(n_forecasts = n(),
  p_max = n_forecasts / max_forecasts * 100)
model_forecasts <- round(summary(model_forecasts$p_max), 0)
# Model structures
structures <- scores |>
  select(Model, Method, agreement) |>
  distinct()
structure_count <- structures |>
  group_by(Method) |>
  summarise(agreement = sum(agreement),
            n = n(),
            p = agreement / n)
# Named lookups so the text below tracks the data rather than hardcoding counts
n_by_method <- setNames(structure_count$n, as.character(structure_count$Method))
n_disagree <- sum(!structures$agreement)
p_disagree <- round(n_disagree / nrow(structures) * 100)
p_by_method <- scores |>
  count(Method) |>
  mutate(p = round(n / sum(n) * 100)) |>
  (\(x) setNames(x$p, as.character(x$Method)))()
# Semi-mechanistic models that drew at least one statistical vote
semi_mech_votes <- classify_models(return_majority = FALSE) |>
  filter(model %in% structures$Model[structures$Method == "Semi-mechanistic"])
n_semi_mech_split <- n_distinct(
  semi_mech_votes$model[semi_mech_votes$classification == "Statistical"]
)

# Country targets
targets <- scores |>
  select(Model, CountryTargets) |>
  distinct() |>
  pull(CountryTargets) |>
  table()
targets <- targets[targets > 0]

single_countries <- scores |>
group_by(epi_target) |>
  filter(CountryTargets == "Single-country") |>
  select(Model, Location) |>
  distinct() |>
  pull(Location) |>
  table()
single_countries <- rev(sort(single_countries[single_countries > 0]))

multi_countries <- scores |>
group_by(epi_target) |>
  filter(CountryTargets == "Multi-country") |>
  select(Model, Location) |>
  distinct() |>
  pull(Location) |>
  table()
multi_countries <- rev(sort(multi_countries[multi_countries > 0]))

model_targets <- table_targets(scores)
multi_targets <- filter(model_targets, CountryTargets == "Multi-country")|>
mutate(diff = max_targets - min_targets)

# Table 1
print_table1(scores)
```

Table 1 summarises model characteristics by structure and geographic target selection; eligibility criteria and exclusion counts are in the STROBE flow diagram (Supplementary Figure S1).
Forecasters could participate in any of 26,624 available targets (104 weeks, 32 countries, 4 horizons, 2 epidemiological outcomes).
Median participation was `r model_forecasts[["Median"]]`% of available targets (IQR `r model_forecasts[["1st Qu."]]`-`r model_forecasts[["3rd Qu."]]`%, range `r model_forecasts[["Min."]]` to `r model_forecasts[["Max."]]`).

We classified `r n_by_method[["Mechanistic"]]` models as mechanistic, `r n_by_method[["Statistical"]]` as statistical, `r n_by_method[["Semi-mechanistic"]]` as semi-mechanistic, `r n_by_method[["Agent-based"]]` as agent-based and `r n_by_method[["Judgement"]]` as human judgement (Table 1; Supplementary Table S2).
Raters disagreed on `r n_disagree` (`r p_disagree`%) classifications, most often for semi-mechanistic models, `r n_semi_mech_split` of `r n_by_method[["Semi-mechanistic"]]` of which drew at least one vote for a statistical structure.
Manual review retained the majority decision in every case.
Statistical models contributed the largest share of forecasts (`r p_by_method[["Statistical"]]`%), partly because the Hub baseline model submitted for every country in almost every week; mechanistic and semi-mechanistic models contributed `r p_by_method[["Mechanistic"]]`% and `r p_by_method[["Semi-mechanistic"]]`%, and agent-based and judgement models under 2% each.

`r targets[["Single-country"]]` models forecast for a single country and `r targets[["Multi-country"]]` for multiple countries, of which only `r nrow(multi_targets |> filter(consistent))` kept the same number of countries throughout (Supplement).

#### Forecast performance

Unadjusted performance varied over space and time, worsening through summer 2021 as the Delta variant emerged.
Agent-based and judgement models appeared to outperform mechanistic and statistical approaches, and single-country models to outperform multi-country models, particularly for deaths from spring 2022.
Model structure was itself distributed unevenly across geographic and epidemiological targets, so these crude comparisons are unreliable.

```{r}
#| label: fig-error-vs-obs
#| fig-cap: >-
#|   Forecast error scales with the magnitude of the observed outcome.
#|   Weighted interval score (WIS) on the natural scale is plotted against observed incidence for COVID-19 cases and deaths across 32 European countries.
#|   Faint points show individual forecasts; lines show a generalised additive model smooth fit per model-structure group, with shaded 95% confidence intervals.
#|   Both axes are shown on the log scale; lower WIS indicates better performance.
#|   Colour: model structure.
#| fig-height: 4
#| fig-width: 8

plot_error_vs_obs(scores_natural)
```

```{r}
#| label: results-scores-results

results <- readRDS(here("output", "log", "results.rds"))
table_effects <- results$effects |> 
  mutate(upper_97.5_text = if_else(
           upper_97.5 < 0,
           paste0("(", round(upper_97.5, 2), ")"),
           as.character(round(upper_97.5, 2))
         )) |>
  mutate(value_ci = paste0(round(value, 2),
                           " (", round(lower_2.5, 2), "-",
                           upper_97.5_text, ")"),
         # exponentiated: multiplicative ratio vs grand-mean LWIS (1 = average)
         ratio_ci = paste0(round(exp(value), 2),
                           " (", round(exp(lower_2.5), 2), "-",
                           round(exp(upper_97.5), 2), ")"),
         group = paste(model, group, sep = "_")) |>
  column_to_rownames("group")

effects_comp <- results$effects |> 
  pivot_longer(cols = c(value, lower_2.5, upper_97.5, se)) |> 
  pivot_wider(names_from = model) |> 
  mutate(diff = Adjusted - Unadjusted,
  diff_percent = diff / Unadjusted * 100) |> 
  filter(group_var %in% c("Method"))
```

#### Effect of model structure

Partial effects are deviations from the grand mean LWIS under a sum-to-zero constraint, so a negative value indicates better-than-average performance.
Exponentiating gives a multiplicative ratio against the grand mean, where 1 is average performance and a partial effect of −0.1 corresponds to a ratio of `r round(exp(-0.1), 2)`, an LWIS `r round((1 - exp(-0.1)) * 100, digits=0)`% lower than average.
We report ratios with 95% confidence intervals here (Table 2), and raw partial effects in the Supplement.
The LWIS is highly right-skewed, so we used a Tweedie family with a log link; a Gaussian family left strongly skewed residuals, and a Gamma fitted comparably but has no support at zero (Supplementary Figures S2-S3).

We let each structure differ in how well it predicted cases and deaths, rather than assuming one effect shared across both.
Pooling across outcomes, no structural approach dominated (Table 2).
A pooled estimate averages a structure's effects on cases and deaths, so a value near the grand mean can arise either from average performance on both or from relatively good performance on one and poor on the other; we therefore also estimated each structure separately by outcome.
The largest shifts under adjustment were among agent-based and human judgement models, which had appeared better than average unadjusted, and adjustment shrank the standard errors throughout (@fig-plot-coeffs).

```{r}
#| label: results-table2

print_table2(results$effects)
```
```{r}
#| label: fig-plot-coeffs
#| fig-cap: >-
#|   Performance ratio (95% CI) on the LWIS from model structure, before and after adjustment for confounding.
#|   The ratio is the exponentiated partial effect relative to the grand-mean LWIS; a ratio below 1 indicates better-than-average performance.
#|   Adjusted effects account for epidemiological target (cases or deaths), forecast horizon, epidemic trend, dominant variant phase, geographic location, and individual model variation.
#|   Filled symbols: adjusted; open symbols: unadjusted.
#|   Ratios and 95% confidence intervals estimated from a generalised additive mixed model.
#|   LWIS = weighted interval score of log-transformed forecasts and observations.
#| fig-height: 3
#| fig-width: 5

plot_effects(results$effects, variables = c("Method"))
```

```{r}
#| label: results-method-target

mbt <- results$method_by_target
mbt_ratio <- setNames(
  sprintf(
    "%.2f (%.2f-%.2f)",
    exp(mbt$value), exp(mbt$lower_2.5), exp(mbt$upper_97.5)
  ),
  paste(mbt$Method, mbt$Epi_target, sep = "_")
)
# case minus death effect on the log scale; negative indicates a structure
# performed relatively better on cases than on deaths
mbt_diff <- vapply(
  split(mbt, mbt$Method),
  \(x) x$value[x$Epi_target == "Cases"] - x$value[x$Epi_target == "Deaths"],
  numeric(1)
)
mbt_widest <- names(which.max(abs(mbt_diff)))
```

Separated by outcome, differences between structures were larger than the pooled estimates suggested, though no structure-by-outcome estimate was distinguishable from the grand mean (Table 3, @fig-method-target).
Structures differed in the direction of their outcome-specific effects, and these largely offset when averaged.
The widest separation was among `r tolower(mbt_widest)` models, the worst-performing structure for case forecasts and the best for death forecasts (cases ratio `r mbt_ratio[[paste0(mbt_widest, "_Cases")]]`; deaths ratio `r mbt_ratio[[paste0(mbt_widest, "_Deaths")]]`), while their pooled estimate sits close to the grand mean.
Mechanistic and statistical models reversed less sharply in the other direction, performing marginally better on cases than on deaths.
An estimate that averages over outcomes therefore recovers close to zero for most structures, whether or not outcome-specific differences exist.
Human judgement models performed relatively better on cases than on deaths (cases ratio `r mbt_ratio[["Judgement_Cases"]]`; deaths ratio `r mbt_ratio[["Judgement_Deaths"]]`), in the same direction as a previous comparison of human and model-based forecasts in this setting [@bosse2022].
The agent-based interval for death forecasts only marginally spans the grand mean, the widest separation rests on only three contributing models, and both the size of these differences and, for two of the five structures, their direction depend on the choice of error family (Supplement).

```{r}
#| label: results-table3

print_table_method_target(results$method_by_target)
```

```{r}
#| label: fig-method-target
#| fig-cap: >-
#|   Performance ratio (95% CI) by model structure, estimated separately for case and death forecasts.
#|   The ratio is the exponentiated partial effect relative to the grand-mean LWIS; a ratio below 1 indicates better-than-average performance.
#|   Estimates come from a single joint model in which model structure is crossed with the epidemiological outcome.
#|   Colour and shape: epidemiological outcome.
#|   LWIS = weighted interval score of log-transformed forecasts and observations.
#| fig-height: 3
#| fig-width: 6

plot_method_target(results$method_by_target)
```

#### Other drivers of forecast performance

Features of the target explained more variation in performance than features of the model (full reporting in the Supplement).
These were modelled jointly across both outcomes, so each estimate is a single effect shared by case and death forecasts.
Targets were most predictable during stable periods of the epidemic curve (ratio `r table_effects["Adjusted_Stable","ratio_ci"]`) and least during increasing trends (ratio `r table_effects["Adjusted_Increasing","ratio_ci"]`).
Predictability was better in the earlier Alpha and Delta phases (ratios `r table_effects["Adjusted_Alpha","ratio_ci"]` and `r table_effects["Adjusted_Delta","ratio_ci"]`) and worst under Omicron BA.1 (ratio `r table_effects["Adjusted_Omicron-BA.1","ratio_ci"]`).
Death forecasts scored lower than case forecasts (ratio `r table_effects["Adjusted_Deaths","ratio_ci"]`).

Substantial variation between individual models remained unexplained (@fig-plot-models), so model structure and geographic scope do not account for the differences in performance between forecasters.

```{r}
#| label: results-ranks
#| include: false

ranks <- rank_models(results$effects)
rank_summary <- summarise_ranks(ranks)
```

Adjustment nonetheless changed which models appeared best: ranking models by their partial effect before and after adjustment gives a Spearman correlation of `r round(rank_summary$spearman, 2)`, with `r rank_summary$n_moved` of `r rank_summary$n` models moving at least `r rank_summary$threshold` places (Supplement).
Observed performance therefore reflects the difficulty of the targets a model chose as much as the model itself.

```{r}
#| label: fig-plot-models
#| fig-cap: >-
#|   Estimated performance ratio (95% CI) by individual model.
#|   The ratio is the exponentiated partial effect relative to the grand-mean LWIS, where a ratio below 1 indicates better-than-average performance.
#|   Adjusted performance after accounting for model structure, geographic scope, forecast horizon, epidemic trend, variant phase, and location.
#|   Colour: model structure.
#|   Shape: geographic scope (circle = single-country, triangle = multi-country).
#|   Remaining variation represents unexplained heterogeneity.
#| fig-height: 8
#| fig-width: 10

plot_models(results$effects, scores)
```

#### Sensitivity analysis

We repeated the analysis on the natural scale, and under alternative covariate sets and an identity link, all reported in the Supplement.
None changed the partial effects materially; the identity link left them unchanged to two decimal places.

The structure-by-outcome estimates were the exception.
Refitting with a Gaussian family in place of the Tweedie changed their size throughout, and their direction for two of the five structures: the separation between cases and deaths widened for human judgement models, disappeared for agent-based models, and reversed for semi-mechanistic models (Supplement).
A Gaussian family is dominated by the largest scores, which fall disproportionately among case forecasts.
Including or excluding the Hub baseline model made no material difference.
We therefore treat the ordering of structures within an outcome as indicative rather than established.
