```{r}
#| label: results-set-up
#| include: false

library(here)
library(dplyr)
library(readr)
library(tidyr)
library(tibble)
library(purrr)
library(ggplot2)
library(patchwork)
library(kableExtra)
theme_set(theme_classic())
knitr::opts_chunk$set(
  eval = TRUE, echo = FALSE,
  message = FALSE, warning = FALSE
)
source(here("R", "process-data.R"))
source(here("R", "analysis-descriptive.R"))
source(here("R", "plot-model-results.R"))
## to re-run and save model results:
# source(here("R", "analysis-model.R")) 
# model_wis(scoring_scale = "log", output_dir = "output")

scores <- process_data(scoring_scale = "log")
ensemble <- scores |>
  filter(grepl("EuroCOVIDhub-ensemble", Model))
scores <- scores |>
  filter(!grepl("EuroCOVIDhub-", Model))
```

<!---Summary--->
```{r}
#| label: results-participation-text

max_forecasts <- 104 * 2 * 4 * 32 # 104 weeks * 2 epi outcomes * 4 horizons * 32 countries
n_models <- length(unique(scores$Model))
n_forecasts <- nrow(scores)
n_teams <- distinct(scores, Model) |> 
  separate(Model, into = c("team", "model"), sep = "-") |> 
  count(team)
teams_models <- summary(n_teams$n)
```

We evaluated the predictive performance of `r n_models` forecasting models produced by `r nrow(n_teams)` separate modelling teams participating in the European COVID-19 Forecast Hub.
We identified both differential participation and varying performance among models, using the weighted interval score (WIS) as a standardised measure of performance.
We then used a generalised additive model to associate predictive performance with model structure.
We accounted for confounding from forecasters' selection of geographic targets (the number of countries a model targeted), and forecast target difficulty (factors associated with the epidemiological context).

**Forecaster characteristics**

```{r}
#| label: results-describe-structures

# Overall participation
model_forecasts <- scores |>
  group_by(Model) |>
  summarise(n_forecasts = n(),
  p_max = n_forecasts / max_forecasts * 100)
model_forecasts <- round(summary(model_forecasts$p_max), 0)
# Model structures
structures <- scores |>
  select(Model, Method, agreement) |>
  distinct()
structure_count <- structures |>
  group_by(Method) |>
  summarise(agreement = sum(agreement),
            n = n(),
            p = agreement / n)

# Country targets
targets <- scores |>
  select(Model, CountryTargets) |>
  distinct() |>
  pull(CountryTargets) |>
  table()
targets <- targets[targets > 0]

single_countries <- scores |>
group_by(epi_target) |>
  filter(CountryTargets == "Single-country") |>
  select(Model, Location) |>
  distinct() |>
  pull(Location) |>
  table()
single_countries <- rev(sort(single_countries[single_countries > 0]))

multi_countries <- scores |>
group_by(epi_target) |>
  filter(CountryTargets == "Multi-country") |>
  select(Model, Location) |>
  distinct() |>
  pull(Location) |>
  table()
multi_countries <- rev(sort(multi_countries[multi_countries > 0]))

model_targets <- table_targets(scores)
multi_targets <- filter(model_targets, CountryTargets == "Multi-country")|>
mutate(diff = max_targets - min_targets)

# Table 1
print_table1(scores)
```

We summarised `r n_models` model characteristics by model structure and geographic target selection (Table 1).
We included all participant forecasts for any target; full eligibility criteria and exclusion counts are shown in the STROBE flow diagram (Supplementary Figure S1).
Over the study period, forecasters could participate in any of 26,624 available forecast targets (comprising 104 weeks, 32 countries, 4 horizons, and 2 epidemiological outcomes).
Median forecaster participation was `r model_forecasts[["Median"]]`% (IQR: `r model_forecasts[["1st Qu."]]`-`r model_forecasts[["3rd Qu."]]`%, ranging from `r model_forecasts[["Min."]]` to `r model_forecasts[["Max."]]`)
across available targets, reflecting substantial variation in participation across countries and time.

We classified 17 models as using mechanistic structure, 12 as statistical, 12 as semi-mechanistic, 3 as agent-based and 3 that used human judgement forecasting (Table 1; Supplementary Table S2).
Raters disagreed on 17 (36%) model classifications.
The majority of 2/3 was used as the final classification, with additional manual review which in all cases retained the majority decision.
Raters most often disagreed when classifying semi-mechanistic models, with 8 out of 12 receiving one or more votes for a statistical model structure.
Under this classification, approximately one-third of all forecast predictions were from mechanistic, semi-mechanistic, and statistical models each.
Agent-based and judgement models provided fewer forecasts, representing only 1-2% of forecasts.

We explored geographic target selection among our binary classification of whether models forecast for only one country (`r targets[["Single-country"]]` models), or multiple countries (`r length(single_countries)` models).
The most common country targets were Germany (`r single_countries[["DE"]]` single-country models), Poland (`r single_countries[["PL"]]`), Italy (`r single_countries[["IT"]]`), Spain (`r single_countries[["ES"]]`), and Czechia (`r single_countries[["CZ"]]`). One additional single-country model targeted Slovenia (`r single_countries[["SI"]]`).
With the exception of Slovenia, these were also the most frequent countries most frequently targeted by multi-country models. However, we noted that among `r targets[["Multi-country"]]` multi-country models, only `r nrow(multi_targets |> filter(consistent))` models consistently forecast for the same number of countries throughout the entire study period.

**Descriptive forecast performance**

```{r}
#| label: results-scores-results

results <- readRDS(here("output", "log", "results.rds"))
table_effects <- results$effects |> 
  mutate(upper_97.5_text = if_else(
           upper_97.5 < 0,
           paste0("(", round(upper_97.5, 2), ")"),
           as.character(round(upper_97.5, 2))
         )) |>
  mutate(value_ci = paste0(round(value, 2), 
                           " (", round(lower_2.5, 2), "-", 
                           upper_97.5_text, ")"),
         group = paste(epi_target, model, group, sep = "_")) |>
  column_to_rownames("group")

effects_comp <- results$effects |> 
  pivot_longer(cols = c(value, lower_2.5, upper_97.5, se)) |> 
  pivot_wider(names_from = model) |> 
  mutate(diff = Adjusted - Unadjusted) |> 
  filter(group_var %in% c("Method", "CountryTargets")) |> 
  select(-c("effect", "Unadjusted"))
```

```{r}
#| label: fig-scores-over-time
#| fig-cap: "Predictive accuracy of forecasting models for COVID-19 cases and deaths across 32 European countries over time. Forecast performance is shown as the mean weighted interval score (WIS) on the log scale; lower scores indicate better performance. WIS was computed on log-transformed incidence per 100,000 population, summarised across 32 countries and 1–4 week horizons. Rows: single- vs multi-country models. Columns: cases vs deaths. Colour: model structure. Uncertainty intervals omitted for clarity (see Supplementary Figure S4)."
#| fig-height: 8
#| fig-width: 6

# plot_over_time(scores = scores, ensemble = ensemble, show_uncertainty = TRUE)
```

We descriptively summarised forecast performance across the study sample.
@fig-scores-over-time shows mean WIS over time, stratified by geographic scope (rows) and epidemiological target (columns), with colour indicating model structure.

Unadjusted performance varied across the entire cohort of models over time, and between model structures.
All models saw worsening performance during summer 2021, coinciding with emergence of the Delta variant.
We noted apparent differences between agent-based and judgement models compared to mechanistic and statistical approaches; and single-country models appearing to outperform multi-country models in unadjusted comparisons, particularly for death outcomes from spring 2022.
However, model structure differed substantially between geographic target selection (table 1), making crude unadjusted comparisons unreliable.

**Adjusted estimates of forecast performance**

We fitted a generalised additive mixed model to give adjusted estimates of the partial effect of model structure, while controlling for target strategy and other sources of variation.
In this structure, all partial effects are deviations from the grand mean WIS under sum-to-zero constraints, meaning that a negative value indicates better-than-average performance.
The WIS was highly right-skewed with respect to all other variables (see Supplementary Figure S2), and we used a log-link to account for this.
We report effects on the log-WIS scale.
Exponentiated values give the multiplicative ratio relative to the grand mean (e.g. a partial effect of −0.1 corresponds to a WIS approximately `r round((1 - exp(-0.1)) * 100)`% lower than average).
Model diagnostic plots showed some deviation from normality in the tails of the residual distribution (Supplementary Figure S3).

<!--- Main results: by model structure --->
```{r}
#| label: fig-plot-coeffs
#| fig-cap: "Partial effect (95% CI) on the weighted interval score from model structure and geographic scope, before and after adjustment for confounding. A lower WIS indicates better forecast performance; effects <0 indicate better-than-average performance. Adjusted effects account for forecast horizon, epidemic trend, dominant variant phase, geographic location, and individual model variation. Filled symbols: adjusted; open symbols: unadjusted. Partial effects and 95% confidence intervals estimated from a generalised additive mixed model. WIS = weighted interval score."
#| fig-height: 3
#| fig-width: 5

plot_effects(results$effects, variables = c("Method", "CountryTargets"))
```

```{r}
#| label: results-table2

print_table2(results$effects)
```

After adjustment for confounding, we found no clear evidence for systematic differences in accuracy between model structures (@fig-plot-coeffs; Table 2).
The largest shifts were among agent-based and human judgement models, which appeared better than average in unadjusted estimates (cases: `r table_effects["Cases_Unadjusted_Agent-based","value_ci"]` and `r table_effects["Cases_Unadjusted_Judgement","value_ci"]`; deaths: `r table_effects["Deaths_Unadjusted_Agent-based","value_ci"]` and `r table_effects["Deaths_Unadjusted_Judgement","value_ci"]`) but showed smaller, less certain differences after adjustment (cases: `r table_effects["Cases_Adjusted_Agent-based","value_ci"]` and `r table_effects["Cases_Adjusted_Judgement","value_ci"]`; deaths: `r table_effects["Deaths_Adjusted_Agent-based","value_ci"]` and `r table_effects["Deaths_Adjusted_Judgement","value_ci"]`).
We continued to see a weak difference among model structures in predictive accuracy for deaths, for which statistical models may have performed slightly better (partial effect `r table_effects["Deaths_Adjusted_Statistical","value_ci"]`, WIS ratio `r round(exp(as.numeric(table_effects["Deaths_Adjusted_Statistical","value"])), 2)`) and semi-mechanistic models worse (`r table_effects["Deaths_Adjusted_Semi-mechanistic","value_ci"]`, ratio `r round(exp(as.numeric(table_effects["Deaths_Adjusted_Semi-mechanistic","value"])), 2)`) than average, with overlapping uncertainty.

Forecast target selection and difficulty may have influenced performance.
In the fully adjusted model, single-country models showed slightly better performance than multi-country models (cases: `r table_effects["Cases_Adjusted_Single-country","value_ci"]` vs `r table_effects["Cases_Adjusted_Multi-country","value_ci"]`; deaths: `r table_effects["Deaths_Adjusted_Single-country","value_ci"]` vs `r table_effects["Deaths_Adjusted_Multi-country","value_ci"]`), after adjustment, although with overlapping uncertainty.
Among all models' overall mean performance, we saw relatively improved predictive performance during stable periods of the outbreak curve (cases: `r table_effects["Cases_Adjusted_Stable","value_ci"]`; deaths: `r table_effects["Deaths_Adjusted_Stable","value_ci"]`), and worse predictive performance when targeting periods of decreasing incidence (cases: `r table_effects["Cases_Adjusted_Decreasing","value_ci"]`; deaths: `r table_effects["Deaths_Adjusted_Decreasing","value_ci"]`).
Forecast performance also varied by dominant variant phase, with worse performance after the first Omicron BA.1 variant (cases: `r table_effects["Cases_Adjusted_Omicron-BA.1","value_ci"]`; deaths: `r table_effects["Deaths_Adjusted_Omicron-BA.1","value_ci"]`), compared to the Alpha (cases: `r table_effects["Cases_Adjusted_Alpha","value_ci"]`; deaths: `r table_effects["Deaths_Adjusted_Alpha","value_ci"]`) and Delta (cases: `r table_effects["Cases_Adjusted_Delta","value_ci"]`; deaths: `r table_effects["Deaths_Adjusted_Delta","value_ci"]`) dominant periods.

We also identified residual confounding, with substantial unexplained variation between individual model performance.
@fig-plot-models shows the adjusted partial effect for each individual model.
This can be interpreted as variation in model-specific performance after accounting for all other factors considered here.
This indicated that our variable selection was not sufficient to fully explain differences in performance.

```{r}
#| label: fig-plot-models
#| fig-cap: "Estimated partial effect size (95% CI) by individual model. Adjusted performance after accounting for model structure, geographic scope, forecast horizon, epidemic trend, variant phase, and location. Colour: model structure. Shape: geographic scope (circle = single-country, triangle = multi-country). Remaining variation represents unexplained heterogeneity."
#| fig-height: 8
#| fig-width: 10

plot_models(results$effects, scores)
```

**Sensitivity analyses**

We repeated this analysis to look at performance on the natural scale (rather than log-transformed predictions and observations), presented in the Supplement.
In developing this work, we considered several alternative model specifications.
We selected an initial set of variables a priori, and changed this selection after updating our view of confounding factors, for example including an effect for dominant variant phase.
We indicate these choices in the Supplement.
Regardless of a priori variable selection, we did not find any meaningful change in partial effect sizes among alternate model specifications.
