```{r}
#| label: results-set-up
#| include: false

library(here)
library(dplyr)
library(readr)
library(tidyr)
library(tibble)
library(purrr)
library(ggplot2)
library(patchwork)
library(kableExtra)
theme_set(theme_classic())
knitr::opts_chunk$set(
  eval = TRUE, echo = FALSE,
  message = FALSE, warning = FALSE
)
source(here("R", "process-data.R"))
source(here("R", "analysis-descriptive.R"))
source(here("R", "plot-model-results.R"))
## to re-run and save model results:
# source(here("R", "analysis-model.R")) 
# model_wis(scoring_scale = "log", output_dir = "output")

scores <- process_data(scoring_scale = "log")
ensemble <- scores |>
  filter(grepl("EuroCOVIDhub-ensemble", Model))
scores <- scores |>
  filter(!grepl("EuroCOVIDhub-ensemble", Model))

# natural-scale scores for descriptive error-vs-observation figure
scores_natural <- process_data(scoring_scale = "natural") |>
  filter(!grepl("EuroCOVIDhub-ensemble", Model))
```

<!---Summary--->
```{r}
#| label: results-participation-text

max_forecasts <- 104 * 2 * 4 * 32 # 104 weeks * 2 epi outcomes * 4 horizons * 32 countries
n_models <- length(unique(scores$Model))
n_forecasts <- nrow(scores)
n_teams <- distinct(scores, Model) |> 
  separate(Model, into = c("team", "model"), sep = "-") |> 
  count(team)
teams_models <- summary(n_teams$n)
```

We evaluated `r n_models` forecasting models from `r nrow(n_teams)` modelling teams participating in the European COVID-19 Forecast Hub.
Models differed both in which targets they forecast and in how well they performed.
We used a generalised additive mixed model to associate performance with model structure, adjusting for the number of countries a model targeted and for the epidemiological context of each target.

#### Forecaster characteristics

```{r}
#| label: tbl-models
#| tbl-cap: >-
#|   Characteristics of models and forecasts sampled from the European COVID-19
#|   Forecast Hub, March 2021-2023. Models (%) shows the number of models and the
#|   percentage of all included models. Single-country shows models targeting one
#|   country as a fraction and percentage of models in each method group.

# Overall participation
model_forecasts <- scores |>
  group_by(Model) |>
  summarise(n_forecasts = n(),
  p_max = n_forecasts / max_forecasts * 100)
model_forecasts <- round(summary(model_forecasts$p_max), 0)
# Model structures
structures <- scores |>
  select(Model, Method, agreement) |>
  distinct()
structure_count <- structures |>
  group_by(Method) |>
  summarise(agreement = sum(agreement),
            n = n(),
            p = agreement / n)
# Named lookups so the text below tracks the data rather than hardcoding counts
n_by_method <- setNames(structure_count$n, as.character(structure_count$Method))
n_disagree <- sum(!structures$agreement)
p_disagree <- round(n_disagree / nrow(structures) * 100)
p_by_method <- scores |>
  count(Method) |>
  mutate(p = round(n / sum(n) * 100)) |>
  (\(x) setNames(x$p, as.character(x$Method)))()
# Semi-mechanistic models that drew at least one statistical vote
semi_mech_votes <- classify_models(return_majority = FALSE) |>
  filter(model %in% structures$Model[structures$Method == "Semi-mechanistic"])
n_semi_mech_split <- n_distinct(
  semi_mech_votes$model[semi_mech_votes$classification == "Statistical"]
)

# Country targets
targets <- scores |>
  select(Model, CountryTargets) |>
  distinct() |>
  pull(CountryTargets) |>
  table()
targets <- targets[targets > 0]

single_countries <- scores |>
group_by(epi_target) |>
  filter(CountryTargets == "Single-country") |>
  select(Model, Location) |>
  distinct() |>
  pull(Location) |>
  table()
single_countries <- rev(sort(single_countries[single_countries > 0]))

multi_countries <- scores |>
group_by(epi_target) |>
  filter(CountryTargets == "Multi-country") |>
  select(Model, Location) |>
  distinct() |>
  pull(Location) |>
  table()
multi_countries <- rev(sort(multi_countries[multi_countries > 0]))

model_targets <- table_targets(scores)
multi_targets <- filter(model_targets, CountryTargets == "Multi-country")|>
mutate(diff = max_targets - min_targets)

print_table1(scores, caption = NULL)
```

@tbl-models summarises model characteristics by structure and geographic target selection.
Forecasters could participate in any of 26,624 available targets (104 weeks, 32 countries, 4 horizons, 2 epidemiological outcomes).
Median participation was `r model_forecasts[["Median"]]`% of available targets (IQR `r model_forecasts[["1st Qu."]]`-`r model_forecasts[["3rd Qu."]]`%, range `r model_forecasts[["Min."]]` to `r model_forecasts[["Max."]]`).

We classified `r n_by_method[["Mechanistic"]]` models as mechanistic, `r n_by_method[["Statistical"]]` as statistical, `r n_by_method[["Semi-mechanistic"]]` as semi-mechanistic, `r n_by_method[["Agent-based"]]` as agent-based and `r n_by_method[["Judgement"]]` as human judgement (@tbl-models; Supplement).
Raters disagreed on `r n_disagree` (`r p_disagree`%) classifications, most often for semi-mechanistic models, `r n_semi_mech_split` of `r n_by_method[["Semi-mechanistic"]]` of which drew at least one vote for a statistical structure.
Statistical models contributed the largest share of forecasts (`r p_by_method[["Statistical"]]`%), partly because the Hub baseline model submitted for every country in almost every week; mechanistic and semi-mechanistic models contributed `r p_by_method[["Mechanistic"]]`% and `r p_by_method[["Semi-mechanistic"]]`%, and agent-based and judgement models under 2% each.

`r targets[["Single-country"]]` models forecast for a single country and `r targets[["Multi-country"]]` for multiple countries, of which only `r nrow(multi_targets |> filter(consistent))` kept the same number of countries throughout (Supplement).

#### Forecast performance

```{r}
#| label: results-describe-performance

# Median LWIS by outcome and by variant phase, to describe unadjusted
# performance without fitting anything
median_by_target <- scores |>
  group_by(epi_target) |>
  summarise(median = round(median(wis, na.rm = TRUE), 2)) |>
  (\(x) setNames(x$median, x$epi_target))()
median_by_phase <- scores |>
  group_by(VariantPhase) |>
  summarise(median = median(wis, na.rm = TRUE)) |>
  arrange(median)
phase_best <- as.character(median_by_phase$VariantPhase[1])
phase_worst <- as.character(median_by_phase$VariantPhase[nrow(median_by_phase)])
```

Unadjusted performance varied with both the outcome forecast and the epidemic context.
Case forecasts scored a median LWIS of `r median_by_target[["Cases"]]` against `r median_by_target[["Deaths"]]` for deaths, and scores tracked the magnitude of the outcome being predicted (@fig-error-vs-obs).
Across variant phases, forecasts were most accurate during the `r phase_best` phase and least accurate during `r phase_worst`, the period of steepest change in reported incidence.

In aggregate, agent-based and judgement models appeared to outperform mechanistic and statistical approaches, particularly for deaths from spring 2022.
Model structure was itself distributed unevenly across geographic and epidemiological targets, so these crude comparisons are unreliable.

```{r}
#| label: results-sparsity

# How thin the sample becomes once cross-classified by the categorical
# covariates, i.e. what a stratified analysis would have to work with
strata_vars <- c("epi_target", "Horizon", "Location", "Trend",
                 "VariantPhase", "CountryTargets", "Method")
strata_possible <- prod(sapply(scores[strata_vars], n_distinct))
strata_obs <- count(scores, across(all_of(strata_vars)))
strata_filled <- nrow(strata_obs)
strata_median <- median(strata_obs$n)
strata_thin <- round(mean(strata_obs$n < 5) * 100)
```

Adjusting for these differences by stratification would leave little to work with.
Cross-classifying forecasts by the categorical covariates alone gives `r format(strata_possible, big.mark = ",")` possible strata, of which `r format(strata_filled, big.mark = ",")` contain any forecast.
The median non-empty stratum holds `r strata_median` forecasts and `r strata_thin`% hold fewer than five, and incidence, a continuous covariate, cannot be stratified without arbitrary binning.

```{r}
#| label: fig-error-vs-obs
#| fig-cap: >-
#|   Forecast error scales with the magnitude of the observed outcome.
#|   Weighted interval score (WIS) on the natural scale is plotted against observed incidence for COVID-19 cases and deaths across 32 European countries.
#|   Faint points show individual forecasts; lines show a generalised additive model smooth fit per model-structure group, with shaded 95% confidence intervals.
#|   Both axes are shown on the log scale; lower WIS indicates better performance.
#|   Colour: model structure.
#| fig-height: 4
#| fig-width: 8

plot_error_vs_obs(scores_natural)
```

```{r}
#| label: results-scores-results

results <- readRDS(here("output", "log", "results.rds"))
table_effects <- results$effects |> 
  mutate(upper_97.5_text = if_else(
           upper_97.5 < 0,
           paste0("(", round(upper_97.5, 2), ")"),
           as.character(round(upper_97.5, 2))
         )) |>
  mutate(value_ci = paste0(round(value, 2),
                           " (", round(lower_2.5, 2), "-",
                           upper_97.5_text, ")"),
         # exponentiated: multiplicative ratio vs grand-mean LWIS (1 = average)
         ratio_ci = paste0(round(exp(value), 2),
                           " (", round(exp(lower_2.5), 2), "-",
                           round(exp(upper_97.5), 2), ")"),
         group = paste(model, group, sep = "_")) |>
  column_to_rownames("group")

effects_comp <- results$effects |> 
  pivot_longer(cols = c(value, lower_2.5, upper_97.5, se)) |> 
  pivot_wider(names_from = model) |> 
  mutate(diff = Adjusted - Unadjusted,
  diff_percent = diff / Unadjusted * 100) |> 
  filter(group_var %in% c("Method"))
```

#### Effect of model structure

We report effects as a ratio against the grand-mean LWIS, where 1 is average performance and 0.9 an LWIS 10% lower than average, with raw partial effects in the Supplement.

No structural approach dominated (@tbl-structure, @fig-structure-effects A).
The largest shifts under adjustment were among agent-based and human judgement models, which had appeared to perform better than average unadjusted, and adjustment shrank the standard errors throughout.
A pooled estimate averages a structure's effects on cases and deaths, so a value near the grand mean can arise either from average performance on both, or from relatively good performance on one and poor on the other.

```{r}
#| label: tbl-structure
#| tbl-cap: >-
#|   Partial effects of model structure on the performance of COVID-19 forecasts
#|   (LWIS, the weighted interval score of log-transformed forecasts and observations),
#|   from univariate (unadjusted) and joint (adjusted) generalised additive mixed models.
#|   Effects are deviations from the grand mean under a sum-to-zero constraint, expressed
#|   as the exponentiated partial effect: a multiplicative ratio relative to the grand-mean
#|   LWIS, where a ratio below 1 indicates better-than-average performance.
#|   Raw partial effects on the log scale are reported in the Supplement.
#|   95% CI = 95% confidence interval.

print_table2(results$effects, caption = NULL)
```
```{r}
#| label: results-method-target

mbt <- results$method_by_target
mbt_ratio <- setNames(
  sprintf(
    "%.2f (%.2f-%.2f)",
    exp(mbt$value), exp(mbt$lower_2.5), exp(mbt$upper_97.5)
  ),
  paste(mbt$Method, mbt$Epi_target, sep = "_")
)
# case minus death effect on the log scale; negative indicates a structure
# performed relatively better on cases than on deaths
mbt_diff <- vapply(
  split(mbt, mbt$Method),
  \(x) x$value[x$Epi_target == "Cases"] - x$value[x$Epi_target == "Deaths"],
  numeric(1)
)
mbt_widest <- names(which.max(abs(mbt_diff)))
```

Separated by outcome, differences between structures were larger than the pooled estimates suggested, though no structure-by-outcome estimate was distinguishable from the grand mean (@fig-structure-effects B; Supplement).
Model structures differed in the direction of their outcome-specific effects, and these largely offset when averaged.
The widest separation was among `r tolower(mbt_widest)` models, the worst-performing structure for case forecasts and the best for death forecasts (cases ratio `r mbt_ratio[[paste0(mbt_widest, "_Cases")]]`; deaths ratio `r mbt_ratio[[paste0(mbt_widest, "_Deaths")]]`), while their pooled estimate sits close to the grand mean.
Mechanistic and statistical models reversed less sharply in the other direction, performing marginally better on cases than on deaths.
An estimate that averages over outcomes therefore recovers close to zero for most structures, whether or not outcome-specific differences exist.
Human judgement models performed relatively better on cases than on deaths (cases ratio `r mbt_ratio[["Judgement_Cases"]]`; deaths ratio `r mbt_ratio[["Judgement_Deaths"]]`), in the same direction as a previous comparison of human and model-based forecasts in this setting [@bosse2022].
The agent-based interval for death forecasts only marginally spans the grand mean, the widest separation rests on only three contributing models, and both the size of these differences and, for two of the five structures, their direction depend on the choice of error family (Supplement).

```{r}
#| label: fig-structure-effects
#| fig-cap: >-
#|   No model structure performed differently from the average once the difficulty of
#|   its targets was accounted for, and structures that predicted one outcome relatively
#|   well predicted the other relatively poorly, so the difference cancels when pooled.
#|   Performance ratio (95% CI) by model structure, pooled across epidemiological
#|   outcomes (A) and estimated separately for case and death forecasts (B).
#|   The ratio is the exponentiated partial effect relative to the grand-mean LWIS;
#|   a ratio below 1 indicates better-than-average performance.
#|   In A, filled symbols are adjusted and open symbols unadjusted estimates.
#|   In B, colour and shape give the epidemiological outcome.
#|   Both panels come from a single joint model in which model structure is crossed
#|   with the epidemiological outcome; note the panels use different scales.
#|   LWIS = weighted interval score of log-transformed forecasts and observations.
#| fig-height: 3.5
#| fig-width: 9

plot_structure_effects(results$effects, results$method_by_target)
```

#### Other drivers of forecast performance

Features of the target explained more variation in performance than features of the model, and were modelled jointly across both outcomes, so each estimate is a single effect shared by case and death forecasts (full reporting in the Supplement).
Targets were most predictable during stable periods of the epidemic curve (ratio `r table_effects["Adjusted_Stable","ratio_ci"]`) and least during increasing trends (ratio `r table_effects["Adjusted_Increasing","ratio_ci"]`).
Predictability was better in the earlier Alpha and Delta phases (ratios `r table_effects["Adjusted_Alpha","ratio_ci"]` and `r table_effects["Adjusted_Delta","ratio_ci"]`) and worst under Omicron BA.1 (ratio `r table_effects["Adjusted_Omicron-BA.1","ratio_ci"]`), and death forecasts scored lower than case forecasts (ratio `r table_effects["Adjusted_Deaths","ratio_ci"]`).

Turning to the forecasters, the covariates describing them left most of the difference between models unexplained.
Substantial variation between individual models remained after adjustment (@fig-model-variation A), so model structure and geographic scope do not account for how differently forecasters performed.

```{r}
#| label: results-ranks
#| include: false

ranks <- rank_models(results$effects)
rank_summary <- summarise_ranks(ranks)
```

Adjustment also changed which models appeared best.
Ranking models by their partial effect before and after adjustment gives a Spearman correlation of `r round(rank_summary$spearman, 2)`, with `r rank_summary$n_moved` of `r rank_summary$n` models moving at least `r rank_summary$threshold` places (@fig-model-variation B).
Observed performance therefore reflects the difficulty of the targets a model chose as much as the model itself.

```{r}
#| label: fig-model-variation
#| fig-cap: >-
#|   Individual models differ widely in performance after adjustment, and adjusting for
#|   target difficulty reorders them substantially, so which models look best depends on
#|   how the evaluation is designed.
#|   (A) Adjusted performance ratio (95% CI) for each model, ordered by effect.
#|   The ratio is the exponentiated partial effect relative to the grand-mean LWIS,
#|   where a ratio below 1 indicates better-than-average performance, adjusted for model
#|   structure, geographic scope, forecast horizon, epidemic trend, variant phase and location.
#|   (B) Rank of each model before and against after adjustment, with the diagonal marking
#|   no change in rank; distance from it is how far a model moved.
#|   Colour: model structure. Shape in A: geographic scope.
#|   LWIS = weighted interval score of log-transformed forecasts and observations.
#| fig-height: 8
#| fig-width: 11

plot_model_variation(results$effects, scores, ranks = ranks)
```

#### Sensitivity analysis

Repeating the analysis on the natural scale of scores, and under alternative covariate sets, changed no estimate materially (Supplement).
The exception was the choice of error family: refitting with a Gaussian in place of the Tweedie changed the size of every structure-by-outcome estimate and the direction of two of the five, widening the case-death difference for human judgement models, removing it for agent-based models, and reversing it for semi-mechanistic models.
A Gaussian family is dominated by the largest scores, which fall disproportionately among case forecasts.
