```{r}
#| label: results-set-up
#| include: false

library(here)
library(dplyr)
library(readr)
library(tidyr)
library(tibble)
library(purrr)
library(ggplot2)
library(patchwork)
library(kableExtra)
theme_set(theme_classic())
knitr::opts_chunk$set(
  eval = TRUE, echo = FALSE,
  message = FALSE, warning = FALSE
)
source(here("R", "process-data.R"))
source(here("R", "analysis-descriptive.R"))
source(here("R", "plot-model-results.R"))
## to re-run and save model results:
# source(here("R", "analysis-model.R")) 
# model_wis(scoring_scale = "log", output_dir = "output")

scores <- process_data(scoring_scale = "log")
ensemble <- scores |>
  filter(grepl("EuroCOVIDhub-ensemble", Model))
scores <- scores |>
  filter(!grepl("EuroCOVIDhub-ensemble", Model))

# natural-scale scores for descriptive error-vs-observation figure
scores_natural <- process_data(scoring_scale = "natural") |>
  filter(!grepl("EuroCOVIDhub-ensemble", Model))
```

<!---Summary--->
```{r}
#| label: results-participation-text

max_forecasts <- 104 * 2 * 4 * 32 # 104 weeks * 2 epi outcomes * 4 horizons * 32 countries
n_models <- length(unique(scores$Model))
n_forecasts <- nrow(scores)
n_teams <- distinct(scores, Model) |> 
  separate(Model, into = c("team", "model"), sep = "-") |> 
  count(team)
teams_models <- summary(n_teams$n)
```

We evaluated the predictive performance of `r n_models` forecasting models produced by `r nrow(n_teams)` separate modelling teams participating in the European COVID-19 Forecast Hub.
We identified both differential participation and varying performance among models, using the weighted interval score (WIS) as a standardised measure of performance.
We then used a generalised additive mixed model (GAMM) to associate predictive performance with model structure.
We accounted for confounding from forecasters' selection of geographic targets (the number of countries a model targeted), and forecast target difficulty (factors associated with the epidemiological context).

**Forecaster characteristics**

```{r}
#| label: results-describe-structures

# Overall participation
model_forecasts <- scores |>
  group_by(Model) |>
  summarise(n_forecasts = n(),
  p_max = n_forecasts / max_forecasts * 100)
model_forecasts <- round(summary(model_forecasts$p_max), 0)
# Model structures
structures <- scores |>
  select(Model, Method, agreement) |>
  distinct()
structure_count <- structures |>
  group_by(Method) |>
  summarise(agreement = sum(agreement),
            n = n(),
            p = agreement / n)
# Named lookups so the text below tracks the data rather than hardcoding counts
n_by_method <- setNames(structure_count$n, as.character(structure_count$Method))
n_disagree <- sum(!structures$agreement)
p_disagree <- round(n_disagree / nrow(structures) * 100)
p_by_method <- scores |>
  count(Method) |>
  mutate(p = round(n / sum(n) * 100)) |>
  (\(x) setNames(x$p, as.character(x$Method)))()
# Semi-mechanistic models that drew at least one statistical vote
semi_mech_votes <- classify_models(return_majority = FALSE) |>
  filter(model %in% structures$Model[structures$Method == "Semi-mechanistic"])
n_semi_mech_split <- n_distinct(
  semi_mech_votes$model[semi_mech_votes$classification == "Statistical"]
)

# Country targets
targets <- scores |>
  select(Model, CountryTargets) |>
  distinct() |>
  pull(CountryTargets) |>
  table()
targets <- targets[targets > 0]

single_countries <- scores |>
group_by(epi_target) |>
  filter(CountryTargets == "Single-country") |>
  select(Model, Location) |>
  distinct() |>
  pull(Location) |>
  table()
single_countries <- rev(sort(single_countries[single_countries > 0]))

multi_countries <- scores |>
group_by(epi_target) |>
  filter(CountryTargets == "Multi-country") |>
  select(Model, Location) |>
  distinct() |>
  pull(Location) |>
  table()
multi_countries <- rev(sort(multi_countries[multi_countries > 0]))

model_targets <- table_targets(scores)
multi_targets <- filter(model_targets, CountryTargets == "Multi-country")|>
mutate(diff = max_targets - min_targets)

# Table 1
print_table1(scores)
```

We summarised `r n_models` model characteristics by model structure and geographic target selection (Table 1).
We included all participant forecasts for any target; full eligibility criteria and exclusion counts are shown in the STROBE flow diagram (Supplementary Figure S1).
Over the study period, forecasters could participate in any of 26,624 available forecast targets (comprising 104 weeks, 32 countries, 4 horizons, and 2 epidemiological outcomes).
Median forecaster participation was `r model_forecasts[["Median"]]`% (IQR: `r model_forecasts[["1st Qu."]]`-`r model_forecasts[["3rd Qu."]]`%, ranging from `r model_forecasts[["Min."]]` to `r model_forecasts[["Max."]]`)
across available targets, reflecting substantial variation in participation across countries and time.

We classified `r n_by_method[["Mechanistic"]]` models as using mechanistic structure, `r n_by_method[["Statistical"]]` as statistical, `r n_by_method[["Semi-mechanistic"]]` as semi-mechanistic, `r n_by_method[["Agent-based"]]` as agent-based and `r n_by_method[["Judgement"]]` that used human judgement forecasting (Table 1; Supplementary Table S2).
Raters disagreed on `r n_disagree` (`r p_disagree`%) model classifications.
The majority of 2/3 was used as the final classification, with additional manual review which in all cases retained the majority decision.
Raters most often disagreed when classifying semi-mechanistic models, with `r n_semi_mech_split` out of `r n_by_method[["Semi-mechanistic"]]` receiving one or more votes for a statistical model structure.
Under the final classification, statistical models contributed the largest share of forecast predictions (`r p_by_method[["Statistical"]]`%), partly because the Hub baseline model submitted for every country in almost every week.
Mechanistic and semi-mechanistic models contributed `r p_by_method[["Mechanistic"]]`% and `r p_by_method[["Semi-mechanistic"]]`% respectively.
Agent-based and judgement models provided fewer forecasts, each representing under 2% of forecasts.

We classified `r targets[["Single-country"]]` models as forecasting for a single country and `r targets[["Multi-country"]]` as forecasting for multiple countries; the countries targeted are detailed in the Supplement.
Among the `r targets[["Multi-country"]]` multi-country models, only `r nrow(multi_targets |> filter(consistent))` consistently forecast for the same number of countries throughout the study period.

**Forecast performance**

We descriptively summarised forecast performance across the study sample.
Unadjusted performance varied across the entire cohort of models over space and time.
We saw consistently worsening performance during summer 2021, coinciding with emergence of the Delta variant.
By model structure, we noted apparent improved performance from agent-based and judgement models, compared to mechanistic and statistical approaches; and single-country models appeared to outperform multi-country models, particularly for death outcomes from spring 2022.
However, model structure differed substantially between geographic targets and across features of the epidemiological target, making crude unadjusted comparisons unreliable.

```{r}
#| label: fig-error-vs-obs
#| fig-cap: >-
#|   Forecast error scales with the magnitude of the observed outcome.
#|   Weighted interval score (WIS) on the natural scale is plotted against observed incidence for COVID-19 cases and deaths across 32 European countries.
#|   Faint points show individual forecasts; lines show a generalised additive model smooth fit per model-structure group, with shaded 95% confidence intervals.
#|   Both axes are shown on the log scale; lower WIS indicates better performance.
#|   Colour: model structure.
#| fig-height: 4
#| fig-width: 8

plot_error_vs_obs(scores_natural)
```

```{r}
#| label: results-scores-results

results <- readRDS(here("output", "log", "results.rds"))
table_effects <- results$effects |> 
  mutate(upper_97.5_text = if_else(
           upper_97.5 < 0,
           paste0("(", round(upper_97.5, 2), ")"),
           as.character(round(upper_97.5, 2))
         )) |>
  mutate(value_ci = paste0(round(value, 2),
                           " (", round(lower_2.5, 2), "-",
                           upper_97.5_text, ")"),
         # exponentiated: multiplicative ratio vs grand-mean LWIS (1 = average)
         ratio_ci = paste0(round(exp(value), 2),
                           " (", round(exp(lower_2.5), 2), "-",
                           round(exp(upper_97.5), 2), ")"),
         group = paste(model, group, sep = "_")) |>
  column_to_rownames("group")

effects_comp <- results$effects |> 
  pivot_longer(cols = c(value, lower_2.5, upper_97.5, se)) |> 
  pivot_wider(names_from = model) |> 
  mutate(diff = Adjusted - Unadjusted,
  diff_percent = diff / Unadjusted * 100) |> 
  filter(group_var %in% c("Method"))
```

**Effect of model structure**

We fitted a generalised additive mixed model to give adjusted estimates of the partial effect of model structure, while controlling for varying forecaster target selection and epidemic dynamics between targets.
Partial effects are deviations from the grand mean LWIS under a sum-to-zero constraint, so a negative value indicates better-than-average performance.
The LWIS was highly right-skewed, and we modelled it with a Tweedie family and a log link: the log link models the score multiplicatively, and the Tweedie family accommodates the skew rather than leaving it in the residuals (diagnostics in Supplementary Figures S2-S3).
We compared this against a Gaussian and a Gamma family: the Gaussian left strongly skewed residuals, and the Gamma fitted comparably to the Tweedie but did not converge (Supplement).
Exponentiating a partial effect gives a multiplicative ratio relative to the grand-mean LWIS, where 1 indicates average performance (e.g. a partial effect of −0.1 corresponds to a ratio of `r round(exp(-0.1), 2)`, an LWIS approximately `r round((1 - exp(-0.1)) * 100, digits=0)`% lower than average).
We report the exponentiated ratio (with 95% confidence intervals) in the main text (Table 2), with the raw partial effects on the log scale in the Supplement.

We allowed each model structure to differ in how well it predicted cases and deaths, rather than assuming a single effect shared across both outcomes.
Pooling across outcomes, no single structural approach dominated (Table 2).
A pooled estimate is the average of a structure's effects on case and death forecasts, so a value near the grand mean can arise either because a structure performed averagely on both outcomes or because it performed relatively well on one and relatively poorly on the other.
We therefore also estimated the effect of structure separately for each outcome.
The largest shifts under adjustment were among agent-based and human judgement models, which appeared better than average in unadjusted estimates.
Adjustment also shrank the standard error compared to univariate estimates (@fig-plot-coeffs).

```{r}
#| label: results-table2

print_table2(results$effects)
```
```{r}
#| label: fig-plot-coeffs
#| fig-cap: >-
#|   Performance ratio (95% CI) on the weighted interval score from model structure, before and after adjustment for confounding.
#|   The ratio is the exponentiated partial effect relative to the grand-mean LWIS; a ratio below 1 indicates better-than-average performance.
#|   Adjusted effects account for epidemiological target (cases or deaths), forecast horizon, epidemic trend, dominant variant phase, geographic location, and individual model variation.
#|   Filled symbols: adjusted; open symbols: unadjusted.
#|   Ratios and 95% confidence intervals estimated from a generalised additive mixed model.
#|   LWIS = weighted interval score of log-transformed forecasts and observations.
#| fig-height: 3
#| fig-width: 5

plot_effects(results$effects, variables = c("Method"))
```

```{r}
#| label: results-method-target

mbt <- results$method_by_target
mbt_ratio <- setNames(
  sprintf(
    "%.2f (%.2f-%.2f)",
    exp(mbt$value), exp(mbt$lower_2.5), exp(mbt$upper_97.5)
  ),
  paste(mbt$Method, mbt$Epi_target, sep = "_")
)
# case minus death effect on the log scale; negative indicates a structure
# performed relatively better on cases than on deaths
mbt_diff <- vapply(
  split(mbt, mbt$Method),
  \(x) x$value[x$Epi_target == "Cases"] - x$value[x$Epi_target == "Deaths"],
  numeric(1)
)
mbt_widest <- names(which.max(abs(mbt_diff)))
```

Separated by outcome, differences between structures were larger than the pooled estimates suggested, though no structure-by-outcome estimate was distinguishable from the grand mean (Table 3, @fig-method-target).
Structures differed in the direction of their outcome-specific effects, and these largely offset when averaged.
The widest separation was among `r tolower(mbt_widest)` models, the worst-performing structure for case forecasts and the best for death forecasts (cases ratio `r mbt_ratio[[paste0(mbt_widest, "_Cases")]]`; deaths ratio `r mbt_ratio[[paste0(mbt_widest, "_Deaths")]]`), while their pooled estimate sits close to the grand mean.
Mechanistic and statistical models reversed less sharply in the other direction, performing marginally better on cases than on deaths.
An estimate that averages over outcomes therefore recovers close to zero for most structures, whether or not outcome-specific differences exist.
Human judgement models performed relatively better on cases than on deaths (cases ratio `r mbt_ratio[["Judgement_Cases"]]`; deaths ratio `r mbt_ratio[["Judgement_Deaths"]]`), in the same direction as a previous comparison of human and model-based forecasts in this setting [@bosse2022].
Every interval spans the grand mean, though the agent-based estimate for death forecasts only marginally so.
The widest separation rests on only three contributing models, and the magnitude of these differences depends on the choice of error family (Supplement), so we treat the direction of these contrasts as the interpretable result rather than their size.

```{r}
#| label: results-table3

print_table_method_target(results$method_by_target)
```

```{r}
#| label: fig-method-target
#| fig-cap: >-
#|   Performance ratio (95% CI) by model structure, estimated separately for case and death forecasts.
#|   The ratio is the exponentiated partial effect relative to the grand-mean LWIS; a ratio below 1 indicates better-than-average performance.
#|   Estimates come from a single joint model in which model structure is crossed with the epidemiological outcome.
#|   Colour and shape: epidemiological outcome.
#|   LWIS = weighted interval score of log-transformed forecasts and observations.
#| fig-height: 3
#| fig-width: 6

plot_method_target(results$method_by_target)
```

**Other drivers of forecast performance**

Among our selection of covariates, we found performance was more driven by the target epidemiological features than by forecaster model features (with full reporting in Supplement).
These features were modelled jointly across both epidemiological targets, so each estimate is a single effect shared by case and death forecasts.
Holding all other features constant, the most predictable forecast targets appeared during periods of stability in the epidemic curve, with better predictive performance relative to periods of increasing or decreasing trends (ratio `r table_effects["Adjusted_Stable","ratio_ci"]`).
Forecasting was more difficult during increasing trends (ratio `r table_effects["Adjusted_Increasing","ratio_ci"]`), compared to stable or decreasing periods.

Predictability also varied between phases of variant dominance.
We estimated better predictive performance during the earlier phases dominated by the Alpha and Delta variants (Alpha ratio: `r table_effects["Adjusted_Alpha","ratio_ci"]`; Delta ratio: `r table_effects["Adjusted_Delta","ratio_ci"]`).
We estimated worse performance after Omicron variants became dominant (with worse performance in the first Omicron BA.1 variant, ratio: `r table_effects["Adjusted_Omicron-BA.1","ratio_ci"]`).
Death forecasts achieved a lower LWIS than case forecasts, as a fixed contrast holding all other features constant (ratio: `r table_effects["Adjusted_Deaths","ratio_ci"]`).

We identified residual confounding, with substantial unexplained variation between individual model performance.
@fig-plot-models shows the adjusted partial effect for each individual model.
This can be interpreted as variation in model-specific performance after accounting for all other factors considered here.
This indicated that the variables we selected to account for variation between forecasters (i.e. model structure; one or multiple geographic targets) were not sufficient to fully explain differences in performance.

```{r}
#| label: fig-plot-models
#| fig-cap: >-
#|   Estimated performance ratio (95% CI) by individual model.
#|   The ratio is the exponentiated partial effect relative to the grand-mean LWIS, where a ratio below 1 indicates better-than-average performance.
#|   Adjusted performance after accounting for model structure, geographic scope, forecast horizon, epidemic trend, variant phase, and location.
#|   Colour: model structure.
#|   Shape: geographic scope (circle = single-country, triangle = multi-country).
#|   Remaining variation represents unexplained heterogeneity.
#| fig-height: 8
#| fig-width: 10

plot_models(results$effects, scores)
```

**Sensitivity analysis**

We repeated this analysis to look at performance on the natural scale (rather than log-transformed predictions and observations), presented in the Supplement.
In developing this work, we considered several alternative model specifications.
We selected an initial set of variables a priori, and changed this selection after updating our view of confounding factors, for example including an effect for dominant variant phase.
We indicate these choices in the Supplement.
Regardless of a priori variable selection, we did not find any meaningful change in partial effect sizes among alternate model specifications.
Refitting with an identity link in place of the log link also left all partial effects unchanged to two decimal places.

The structure-by-outcome estimates were the exception, being the one set of estimates sensitive to the choice of error family.
Refitting with a Gaussian family in place of the Tweedie preserved the direction of every contrast, but changed their magnitudes: the separation between cases and deaths widened for human judgement models and disappeared for agent-based models (Supplement).
A Gaussian family is dominated by the largest scores, which fall disproportionately among case forecasts.
Including or excluding the Hub baseline model made no material difference to these estimates.
We therefore treat the ordering of structures within an outcome as indicative rather than established.
