```{r}
#| label: results-set-up
#| include: false

library(here)
library(dplyr)
library(readr)
library(tidyr)
library(tibble)
library(purrr)
library(ggplot2)
library(patchwork)
library(kableExtra)
theme_set(theme_classic())
knitr::opts_chunk$set(
  eval = TRUE, echo = FALSE,
  message = FALSE, warning = FALSE
)
source(here("R", "process-data.R"))
source(here("R", "analysis-descriptive.R"))
source(here("R", "plot-model-results.R"))
## to re-run and save model results:
# source(here("R", "analysis-model.R")) 
# model_wis(scoring_scale = "log", output_dir = "output")

scores <- process_data(scoring_scale = "log")
ensemble <- scores |>
  filter(grepl("EuroCOVIDhub-ensemble", Model))
scores <- scores |>
  filter(!grepl("EuroCOVIDhub-ensemble", Model))

# natural-scale scores for descriptive error-vs-observation figure
scores_natural <- process_data(scoring_scale = "natural") |>
  filter(!grepl("EuroCOVIDhub-ensemble", Model))
```

```{r}
#| label: results-participation-text

max_forecasts <- 104 * 2 * 4 * 32 # 104 weeks * 2 epi outcomes * 4 horizons * 32 countries
n_models <- length(unique(scores$Model))
n_forecasts <- nrow(scores)
n_teams <- distinct(scores, Model) |> 
  separate(Model, into = c("team", "model"), sep = "-") |> 
  count(team)
teams_models <- summary(n_teams$n)
```

We evaluated `r n_models` forecasting models from `r nrow(n_teams)` modelling teams participating in the European COVID-19 Forecast Hub over 2021-2023.
Models differed both in which targets they forecast and in how well they performed.
We used a generalised additive mixed model to associate performance with model structure, adjusting for the number of countries a model targeted and for the epidemiological context of each target.

#### Participation and characteristics

```{r}
#| label: tbl-models
#| tbl-cap: >-
#|   Characteristics of models and forecasts sampled from the European COVID-19
#|   Forecast Hub, March 2021-2023. Models (%) shows the number of models and the
#|   percentage of all included models. Single-country shows models targeting one
#|   country as a fraction and percentage of models in each method group.

# Overall participation
model_forecasts <- scores |>
  group_by(Model) |>
  summarise(n_forecasts = n(),
  p_max = n_forecasts / max_forecasts * 100)
model_forecasts <- round(summary(model_forecasts$p_max), 0)
# Model structures
structures <- scores |>
  select(Model, Method, agreement) |>
  distinct()
structure_count <- structures |>
  group_by(Method) |>
  summarise(agreement = sum(agreement),
            n = n(),
            p = agreement / n)
# Named lookups so the text below tracks the data rather than hardcoding counts
n_by_method <- setNames(structure_count$n, as.character(structure_count$Method))
n_disagree <- sum(!structures$agreement)
p_disagree <- round(n_disagree / nrow(structures) * 100)
p_by_method <- scores |>
  count(Method) |>
  mutate(p = round(n / sum(n) * 100)) |>
  (\(x) setNames(x$p, as.character(x$Method)))()
# Semi-mechanistic models that drew at least one statistical vote
semi_mech_votes <- classify_models(return_majority = FALSE) |>
  filter(model %in% structures$Model[structures$Method == "Semi-mechanistic"])
n_semi_mech_split <- n_distinct(
  semi_mech_votes$model[semi_mech_votes$classification == "Statistical"]
)

# Country targets
targets <- scores |>
  select(Model, CountryTargets) |>
  distinct() |>
  pull(CountryTargets) |>
  table()
targets <- targets[targets > 0]

single_countries <- scores |>
group_by(epi_target) |>
  filter(CountryTargets == "Single-country") |>
  select(Model, Location) |>
  distinct() |>
  pull(Location) |>
  table()
single_countries <- rev(sort(single_countries[single_countries > 0]))

multi_countries <- scores |>
group_by(epi_target) |>
  filter(CountryTargets == "Multi-country") |>
  select(Model, Location) |>
  distinct() |>
  pull(Location) |>
  table()
multi_countries <- rev(sort(multi_countries[multi_countries > 0]))

model_targets <- table_targets(scores)
multi_targets <- filter(model_targets, CountryTargets == "Multi-country")|>
mutate(diff = max_targets - min_targets)

print_table1(scores, caption = NULL)
```

<!--TODO drop geo scope rows, show overall combined column (?with % participation across N targets?) -->

Forecasters could participate in any of 26,624 available targets (104 weeks, 32 countries, 4 horizons, 2 epidemiological outcomes).
Median participation was `r model_forecasts[["Median"]]`% of available targets (IQR `r model_forecasts[["1st Qu."]]`-`r model_forecasts[["3rd Qu."]]`%, range `r model_forecasts[["Min."]]` to `r model_forecasts[["Max."]]`).
<!--TODO add the sparse-matrix numbers here and a ref to the Supplement figure-->

We classified `r n_by_method[["Mechanistic"]]` models as mechanistic, `r n_by_method[["Statistical"]]` as statistical, `r n_by_method[["Semi-mechanistic"]]` as semi-mechanistic, `r n_by_method[["Agent-based"]]` as agent-based and `r n_by_method[["Judgement"]]` as human judgement (@tbl-models; Supplement).
Raters disagreed on `r n_disagree` (`r p_disagree`%) classifications, most often for semi-mechanistic models, `r n_semi_mech_split` of `r n_by_method[["Semi-mechanistic"]]` of which drew at least one vote for a statistical structure.
Statistical models contributed the largest share of predictions (`r p_by_method[["Statistical"]]`%), partly because the Hub baseline model submitted for every country in almost every week; mechanistic and semi-mechanistic models contributed `r p_by_method[["Mechanistic"]]`% and `r p_by_method[["Semi-mechanistic"]]`%, and agent-based and judgement models under 2% each.

#### Forecast performance

```{r}
#| label: fig-error-vs-obs
#| fig-cap: >-
#|   Forecast error scales with the magnitude of the observed outcome.
#|   Weighted interval score (WIS) on the natural scale is plotted against observed incidence for COVID-19 cases and deaths across 32 European countries.
#|   Faint points show individual forecasts; lines show a generalised additive model smooth fit per model-structure group, with shaded 95% confidence intervals.
#|   Both axes are shown on the log scale; lower WIS indicates better performance.
#|   Colour: model structure.
#| fig-height: 4
#| fig-width: 8

plot_error_vs_obs(scores_natural)
```

<!--TODO some very aggregate figure of observed per 100k against performance -->

```{r}
#| label: results-describe-performance

# Median LWIS by outcome and by variant phase, to describe unadjusted
# performance without fitting anything
median_by_target <- scores |>
  group_by(epi_target) |>
  summarise(median = round(median(wis, na.rm = TRUE), 2)) |>
  (\(x) setNames(x$median, x$epi_target))()
median_by_phase <- scores |>
  group_by(VariantPhase) |>
  summarise(median = median(wis, na.rm = TRUE)) |>
  arrange(median)
phase_best <- as.character(median_by_phase$VariantPhase[1])
phase_worst <- as.character(median_by_phase$VariantPhase[nrow(median_by_phase)])
```

<!-- TODO some super aggregated performance measure here, possibly against the Hub ensemble -->Considering crude aggregated performance.
Variation was driven by epidemiological processes over time.
Case forecasts scored a median LWIS of `r median_by_target[["Cases"]]` against `r median_by_target[["Deaths"]]` for deaths, and scores tracked the magnitude of the outcome being predicted (@fig-error-vs-obs).
Across variant phases, forecasts were most accurate during the `r phase_best` phase and least accurate during `r phase_worst`, the period of steepest change in reported incidence.
<!--TODO something about forecast horizon? -->

```{r}
#| label: results-sparsity

# How thin the sample becomes once cross-classified by the categorical
# covariates, i.e. what a stratified analysis would have to work with
strata_vars <- c("epi_target", "Horizon", "Location", "Trend",
                 "VariantPhase", "CountryTargets", "Method")
strata_possible <- prod(sapply(scores[strata_vars], n_distinct))
strata_obs <- count(scores, across(all_of(strata_vars)))
strata_filled <- round(nrow(strata_obs) / strata_possible * 100)
strata_median <- median(strata_obs$n)
strata_thin <- round(mean(strata_obs$n < 5) * 100)
```

#### Effect of model structure

In crude comparison, agent-based and human judgement models appeared to outperform mechanistic and statistical approaches.
However, participation between model structures was unevenly distributed across forecast targets.
Handling this with a stratification approach would produce sparse data, given both the range of targets, and number and type of covariates we wish to adjust for.
To demonstrate, cross-classifying forecasts by the categorical covariates alone (epidemiological outcome, location, epidemic trend, variant phase, number of countries targeted) gives `r format(strata_possible, big.mark = ",")` possible strata, of which `r strata_filled` % contain any forecast.
Among these, each non-empty stratum holds only `r strata_median` forecasts and `r strata_thin`% hold fewer than five forecasts.

```{r}
#| label: results-scores-results

results <- readRDS(here("output", "log", "results.rds"))
table_effects <- results$effects |> 
  mutate(upper_97.5_text = if_else(
           upper_97.5 < 0,
           paste0("(", round(upper_97.5, 2), ")"),
           as.character(round(upper_97.5, 2))
         )) |>
  mutate(value_ci = paste0(round(value, 2),
                           " (", round(lower_2.5, 2), "-",
                           upper_97.5_text, ")"),
         # exponentiated: multiplicative ratio vs grand-mean LWIS (1 = average)
         ratio_ci = paste0(round(exp(value), 2),
                           " (", round(exp(lower_2.5), 2), "-",
                           round(exp(upper_97.5), 2), ")"),
         group = paste(model, group, sep = "_")) |>
  column_to_rownames("group")

effects_comp <- results$effects |> 
  pivot_longer(cols = c(value, lower_2.5, upper_97.5, se)) |> 
  pivot_wider(names_from = model) |> 
  mutate(diff = Adjusted - Unadjusted,
  diff_percent = diff / Unadjusted * 100) |> 
  filter(group_var %in% c("Method"))
```

We report regression results as a ratio against the grand-mean LWIS, where 1 is average performance and 0.9 an LWIS 10% lower than average, with raw partial effects in the Supplement.
After adjustment, no structural approach dominated (@tbl-structure, @fig-structure-effects A).
The largest shifts under adjustment were among agent-based and human judgement models, while adjustment shrank the standard errors throughout.

```{r}
#| label: tbl-structure
#| tbl-cap: >-
#|   Partial effects of model structure on the performance of COVID-19 forecasts
#|   (LWIS, the weighted interval score of log-transformed forecasts and observations),
#|   from univariate (unadjusted) and joint (adjusted) generalised additive mixed models.
#|   Effects are deviations from the grand mean under a sum-to-zero constraint, expressed
#|   as the exponentiated partial effect: a multiplicative ratio relative to the grand-mean
#|   LWIS, where a ratio below 1 indicates better-than-average performance.
#|   Raw partial effects on the log scale are reported in the Supplement.
#|   95% CI = 95% confidence interval.

print_table2(results$effects, caption = NULL)
```

```{r}
#| label: results-method-target

mbt <- results$method_by_target
mbt_ratio <- setNames(
  sprintf(
    "%.2f (%.2f-%.2f)",
    exp(mbt$value), exp(mbt$lower_2.5), exp(mbt$upper_97.5)
  ),
  paste(mbt$Method, mbt$Epi_target, sep = "_")
)
# case minus death effect on the log scale; negative indicates a structure
# performed relatively better on cases than on deaths
mbt_diff <- vapply(
  split(mbt, mbt$Method),
  \(x) x$value[x$Epi_target == "Cases"] - x$value[x$Epi_target == "Deaths"],
  numeric(1)
)
mbt_widest <- names(which.max(abs(mbt_diff)))
```

Differences between model structures also reduced when separated by outcome (@fig-structure-effects B; Supplement).
However, model structures differed in the direction of their outcome-specific effects.
The widest separation was among `r tolower(mbt_widest)` models, which were the worst-performing model structure for case forecasts but the best for death forecasts (cases ratio `r mbt_ratio[[paste0(mbt_widest, "_Cases")]]`; deaths ratio `r mbt_ratio[[paste0(mbt_widest, "_Deaths")]]`), while their pooled estimate sits close to the grand mean.
<!--TODO note: check whether we only have eg 1 ABM forecasting for each target, so this is weak, even though we also include a model-specific effect-->

All other model structures performed more similarly between epidemiological outcomes, performing marginally better on cases than on deaths.
We noted that differences between outcomes largely offset when averaged; an averaged estimate therefore gives close to zero for most structures, whether or not outcome-specific differences exist.

<!--TODO relabel axes to add "Adjusted performance ratio" ie matching text-->

```{r}
#| label: fig-structure-effects
#| fig-cap: >-
#|   No model structure performed differently from the average once the difficulty of
#|   its targets was accounted for; structures that predicted one outcome relatively
#|   well predicted the other relatively poorly, so the difference cancels when pooled.
#|   Performance ratio (95% CI) by model structure, pooled across epidemiological
#|   outcomes (A) and estimated separately for case and death forecasts (B).
#|   The ratio is the exponentiated partial effect relative to the grand-mean LWIS;
#|   a ratio below 1 indicates better-than-average performance.
#|   In A, filled symbols are adjusted and open symbols unadjusted estimates.
#|   In B, colour and shape give the epidemiological outcome.
#|   Both panels come from a single joint model in which model structure is crossed
#|   with the epidemiological outcome; note the panels use different scales.
#|   LWIS = weighted interval score of log-transformed forecasts and observations.
#| fig-height: 3.5
#| fig-width: 9

plot_structure_effects(results$effects, results$method_by_target)
```

#### Other drivers of forecast performance

Target-generating processes explained more variation in performance than features of the forecasting process.
Targets were most predictable during stable periods of the epidemic curve (ratio `r table_effects["Adjusted_Stable","ratio_ci"]`) and least during increasing trends (ratio `r table_effects["Adjusted_Increasing","ratio_ci"]`).
Predictability was better in the earlier Alpha and Delta phases (ratios `r table_effects["Adjusted_Alpha","ratio_ci"]` and `r table_effects["Adjusted_Delta","ratio_ci"]`) and worst under Omicron BA.1 (ratio `r table_effects["Adjusted_Omicron-BA.1","ratio_ci"]`), and death forecasts scored lower than case forecasts (ratio `r table_effects["Adjusted_Deaths","ratio_ci"]`).

Despite some covariates accounting for the forecasting process, most of the difference between individual model performance was left unexplained (@fig-model-variation A).
We accounted for whether models forecast for one or multiple countries, with `r targets[["Single-country"]]` models forecasting for a single country and `r targets[["Multi-country"]]` for multiple countries (of which only `r nrow(multi_targets |> filter(consistent))` kept the same number of countries throughout, see Supplement).
<!--TODO add ratio_ci for geo scope; consider where to report diffs by horizon, probably here-->

```{r}
#| label: results-ranks
#| include: false

ranks <- rank_models(results$effects)
rank_summary <- summarise_ranks(ranks)
```

We compared this remaining variation between individual models in terms of ranked model order.
Ranking models by their partial effect before and after adjustment gives a Spearman correlation of `r round(rank_summary$spearman, 2)`, with `r rank_summary$n_moved` of `r rank_summary$n` models moving at least `r rank_summary$threshold` places (@fig-model-variation B).
<!--TODO is spearman correct here for rank? not wilcoxon or similar?-->

<!--TODO drop the shape on the geom point as it's not the main interest and relabel the individual model labels on the y axis to "Crude rank {overall crude unadjusted rank}", relabel x-axis to "Adjusted performance ratio"-->

```{r}
#| label: fig-model-variation
#| fig-cap: >-
#|   Individual models differ widely in performance after adjustment, and adjusting for
#|   target difficulty reorders them substantially, so which models look best depends on
#|   how the evaluation is designed.
#|   (A) Adjusted performance ratio (95% CI) for each model, ordered by effect.
#|   The ratio is the exponentiated partial effect relative to the grand-mean LWIS,
#|   where a ratio below 1 indicates better-than-average performance, adjusted for model
#|   structure, geographic scope, forecast horizon, epidemic trend, variant phase and location.
#|   (B) Rank of each model before and against after adjustment, with the diagonal marking
#|   no change in rank; distance from it is how far a model moved.
#|   Colour: model structure. Shape in A: geographic scope.
#|   LWIS = weighted interval score of log-transformed forecasts and observations.
#| fig-height: 8
#| fig-width: 11

plot_model_variation(results$effects, scores, ranks = ranks)
```

#### Sensitivity analysis

Repeating the analysis on the natural scale of scores, and under alternative covariate sets, changed no estimate materially (Supplement).
The exception was the choice of error family: refitting with a Gaussian in place of the Tweedie changed the size of every structure-by-outcome estimate and the direction of two of the five.
Specifically, this widened the case-death difference for human judgement models, removed it for agent-based models, and reversed it for semi-mechanistic models.
We explain this where using a Gaussian family is dominated by the largest scores, which fall disproportionately among case forecasts.
