---
title: "Descriptive scores and data checks"
subtitle: "Weighted Interval Score, its components, and the observed values underlying it"
format:
  html:
    toc: true
    toc-depth: 3
    fig-width: 10
    fig-height: 8
execute:
  echo: false
  warning: false
  message: false
  freeze: auto
---

```{r}
#| label: setup

library(here)
library(dplyr)
library(tidyr)
library(readr)
library(purrr)
library(forcats)
library(stringr)
library(ggplot2)
library(plotly)
library(viridisLite)
library(scales)

source(here("R", "process-data.R"))

# Joined scores + model metadata (Method, CountryTargets) for both scales.
# Every model is kept, including the EuroCOVIDhub-* baseline, persistence and
# ensemble reference models that the analysis drops.
# process_data() is the expensive call, so both the scoring views and the
# observed-data checks at the foot of this page read from this one object.
scores_all <- bind_rows(
  log     = process_data("log"),
  natural = process_data("natural"),
  .id = "scale_id"
) |>
  filter(!is.na(wis))

# --- Broken-forecast flag (honest, not dropped) ------------------------------
# ILM-EKF submitted death predictions ~= population (e.g. Greece 2022-08-06,
# 10.76M vs 378 observed). Observed deaths per 100k peak ~32, so death WIS is
# tiny (99.9th percentile ~19); a death WIS > 100 can only come from a wildly
# implausible prediction. Cases legitimately reach WIS in the thousands
# (observed cases per 100k reach ~6,200), so magnitude cannot flag cases — the
# flag is deaths-only.
broken_keys <- scores_all |>
  filter(scale_id == "natural", epi_target == "Deaths", wis > 100) |>
  distinct(Model, Location, target_end_date, epi_target)

scores_all <- scores_all |>
  left_join(mutate(broken_keys, broken = TRUE),
            by = c("Model", "Location", "target_end_date", "epi_target")) |>
  mutate(broken = coalesce(broken, FALSE))

# as.character: Model is a factor; keeping it as one corrupts later c() concats
# (c(factor, "x") coerces the factor to integer codes).
# Flag computed across all horizons so model identity is stable; plots below are
# restricted to the 1-week horizon.
broken_models <- scores_all |> filter(broken) |> distinct(Model) |> pull(Model) |>
  as.character()

# Ensemble marked distinctly from broken models in every model-level label.
ensemble_model <- "EuroCOVIDhub-ensemble"
mark <- function(m) dplyr::case_when(
  m %in% broken_models ~ " ⚠",
  m == ensemble_model  ~ " ★",
  TRUE                 ~ ""
)

# Every scoring plot in this document is restricted to the 1-week-ahead horizon.
# The observed-data checks below use `scores_all` (all horizons) instead.
scores <- scores_all |> filter(Horizon == 1)

# --- Shared aesthetics -------------------------------------------------------
# Fixed, colourblind-safe mapping for the three WIS components, used in EVERY
# stacked view so a colour always means the same component.
comp_levels <- c("dispersion", "overprediction", "underprediction")
comp_cols <- c(
  dispersion      = "#4d4d4d",  # grey  — spread/sharpness penalty
  overprediction  = "#c0392b",  # red   — forecast too high
  underprediction = "#2166ac"   # blue  — forecast too low
)
comp_labs <- c(dispersion = "Dispersion",
               overprediction = "Over-prediction",
               underprediction = "Under-prediction")

# --- WIS accuracy bands ------------------------------------------------------
# Log-scale WIS is in log(inc + 1) units, so a WIS of x is roughly a
# multiplicative error of exp(x); cuts are set at interpretable factors.
log_breaks <- c(-Inf, 0.10, 0.22, 0.41, 0.69, 1.10, Inf)
log_labs   <- c("within ~10%", "~10-25%", "~25-50%", "~50%-2x", "2-3x", ">3x off")
# Natural scale spans five orders of magnitude: decade cuts, shared across
# targets so one colour = one absolute magnitude in every panel.
nat_breaks <- c(-Inf, 0.1, 1, 10, 100, 1000, Inf)
nat_labs   <- c("<0.1", "0.1-1", "1-10", "10-100", "100-1,000", ">1,000")
# Unnamed length-6 palette: scale_fill_manual maps by level order, so the same
# vector serves either label set. Darker = worse.
band_pal <- viridisLite::viridis(6, option = "magma", direction = -1)

# Call after filtering to one scale; `scale_name` is scalar. Not named `scale`:
# the scores data carry a column of that name, which would shadow the argument
# inside a mutate() data mask.
wis_band <- function(wis, scale_name) {
  brk <- if (scale_name == "log") log_breaks else nat_breaks
  lab <- if (scale_name == "log") log_labs   else nat_labs
  cut(wis, brk, lab, right = FALSE)
}

theme_desc <- theme_minimal(base_size = 11) +
  theme(legend.position = "bottom",
        panel.grid.minor = element_blank())

# tag broken models in labels
tag_broken <- function(x) ifelse(x %in% broken_models, paste0(x, " ⚠"), x)

# Stable model numbering for axis labels: rank by overall mean log-scale WIS
# (No. 1 = best overall). Same number used across every view; decode via the
# Model lookup table below.
model_meta <- scores |>
  filter(scale_id == "log") |>
  group_by(Model) |>
  summarise(mean_log_wis = mean(wis), Method = dplyr::first(Method),
            .groups = "drop") |>
  arrange(mean_log_wis) |>
  mutate(No = row_number())

id_num <- setNames(model_meta$No, model_meta$Model)
# axis label = number + marker (⚠ broken, ★ ensemble); full name kept in hover
id_lab <- function(m) paste0(id_num[m], mark(m))

# Long form for stacked component bars
components_long <- function(df) {
  df |>
    pivot_longer(all_of(comp_levels), names_to = "component", values_to = "value") |>
    mutate(component = factor(component, levels = comp_levels))
}
```

Descriptive views of the raw forecast scores: the Weighted Interval Score (WIS) and its additive decomposition into `dispersion + over-prediction + under-prediction`, for 40+ models across two targets and 32 countries.
The final section checks the observed values underlying these scores.

All WIS are computed on incidence per 100,000 population for both natural and log scales.

Note:

- All scoring plots are restricted to the 1-week-ahead horizon.
- Model identifiers use numbers instead of names; these are stable across all views and ranked by overall mean log-scale WIS (1 = best overall).
- The triangle ⚠ marks `r length(broken_models)` models carrying anomalous forecasts, the star ★ marks the ensemble reference model
- The heatmaps are static; the bar charts are interactive, so hover for full model names

```{r}
#| label: helpers

# Mean components per grouping var, ranked by mean WIS. Returns long form.
rank_decomp <- function(df, group) {
  agg <- df |>
    group_by(epi_target, {{ group }}) |>
    summarise(across(all_of(comp_levels), mean),
              wis = mean(wis), n = n(), .groups = "drop")
  agg
}

# Share of forecasts in each accuracy band, per grouping var. Banding is applied
# to raw forecast scores (not to a mean), so extremes affect only their own band.
band_share <- function(df, group, scale_name) {
  df |>
    mutate(band = wis_band(wis, .env$scale_name)) |>
    count(epi_target, {{ group }}, band, .drop = FALSE) |>
    group_by(epi_target, {{ group }}) |>
    mutate(share = n / sum(n)) |>
    ungroup() |>
    filter(!is.na(share))
}

# Stacked 100% bars of band share. `yvar` already ordered by the caller.
band_plot <- function(d, yvar, name_col, ylab) {
  ggplot(d, aes(x = n, y = {{ yvar }}, fill = band,
                text = paste0({{ name_col }}, "<br>", band, "<br>",
                              scales::percent(share, accuracy = 0.1),
                              " (", n, " forecasts)"))) +
    geom_col(position = "fill") +
    facet_wrap(~epi_target) +
    scale_fill_manual(values = band_pal, drop = FALSE, name = NULL) +
    scale_x_continuous(labels = scales::percent) +
    labs(x = ylab, y = NULL) +
    theme_desc
}

stacked_plot <- function(agg, yvar, ylab, name_col) {
  d <- components_long(agg)
  p <- ggplot(d, aes(x = value, y = {{ yvar }}, fill = component,
                     text = paste0({{ name_col }}, "<br>",
                                   comp_labs[as.character(component)],
                                   ": ", round(value, 2)))) +
    geom_col() +
    facet_wrap(~epi_target, scales = "free_x") +
    scale_fill_manual(values = comp_cols, labels = comp_labs, name = NULL) +
    labs(x = ylab, y = NULL) +
    theme_desc
  p
}
```

::: {.callout-note collapse="true"}
## Model lookup {#lookup}

```{r}
#| label: lookup

model_meta |>
  transmute(No,
            Model,
            Structure = as.character(Method),
            Flag = trimws(mark(Model))) |>
  knitr::kable()
```
:::

## Ranked model decomposition {#view1}

Mean WIS per model split into components.

::: {.panel-tabset}
### Log scale

```{r}
#| label: v1-log

agg <- scores |>
  filter(scale_id == "log") |>
  rank_decomp(Model) |>
  mutate(lab = fct_reorder(id_lab(Model), wis, .desc = TRUE))
p <- stacked_plot(agg, lab, "Mean WIS (log scale)", name_col = Model)
ggplotly(p, tooltip = "text") |> layout(legend = list(orientation = "h"))
```

### Natural scale

Note: the ⚠ models' over-prediction bars run off the useful range so the natural scale is dominated by one extreme (ILM-EKF, 2022-08).

```{r}
#| label: v1-nat

agg <- scores |>
  filter(scale_id == "natural") |>
  rank_decomp(Model) |>
  mutate(lab = fct_reorder(id_lab(Model), wis, .desc = TRUE))
p <- stacked_plot(agg, lab, "Mean WIS (natural scale)", name_col = Model)
ggplotly(p, tooltip = "text") |> layout(legend = list(orientation = "h"))
```
:::

## By model structure {#view2}

::: {.panel-tabset}
### Log scale

```{r}
#| label: v2-log

agg <- scores |>
  filter(scale_id == "log", !is.na(Method)) |>
  rank_decomp(Method) |>
  mutate(Method = fct_reorder(Method, wis, .desc = TRUE))
p <- stacked_plot(agg, Method, "Mean WIS (log scale)", name_col = Method)
ggplotly(p, tooltip = "text") |> layout(legend = list(orientation = "h"))
```

### Natural scale

```{r}
#| label: v2-nat

agg <- scores |>
  filter(scale_id == "natural", !is.na(Method)) |>
  rank_decomp(Method) |>
  mutate(Method = fct_reorder(Method, wis, .desc = TRUE))
p <- stacked_plot(agg, Method, "Mean WIS (natural scale)", name_col = Method)
ggplotly(p, tooltip = "text") |> layout(legend = list(orientation = "h"))
```
:::

## Accuracy bands {#bands}

Each forecast is binned into an accuracy band, and the bars show the share of a model's forecasts in each band.

- On the log scale, bands are labelled by relative error: within \~10%, \~10-25%, \~25-50%, \~50%-2x, 2-3x, and >3x off
- On the natural scale bands are decades of WIS per 100,000, shared by both targets, so one colour means one absolute magnitude everywhere

Band shares are more robust to the extreme forecasts, so a model can rank poorly on mean WIS while remaining in a band close to its peers.

::: {.panel-tabset}
### Log scale

```{r}
#| label: bands-log-model

d <- scores |>
  filter(scale_id == "log") |>
  band_share(Model, "log") |>
  mutate(lab = factor(id_lab(Model), levels = rev(id_lab(model_meta$Model))))
p <- band_plot(d, lab, Model, "Share of forecasts (log-scale WIS band)")
ggplotly(p, tooltip = "text") |> layout(legend = list(orientation = "h"))
```

```{r}
#| label: bands-log-method
#| fig-height: 4

d <- scores |>
  filter(scale_id == "log", !is.na(Method)) |>
  band_share(Method, "log")
p <- band_plot(d, Method, Method, "Share of forecasts (log-scale WIS band)")
ggplotly(p, tooltip = "text") |> layout(legend = list(orientation = "h"))
```

### Natural scale

```{r}
#| label: bands-nat-model

d <- scores |>
  filter(scale_id == "natural") |>
  band_share(Model, "natural") |>
  mutate(lab = factor(id_lab(Model), levels = rev(id_lab(model_meta$Model))))
p <- band_plot(d, lab, Model, "Share of forecasts (WIS per 100k band)")
ggplotly(p, tooltip = "text") |> layout(legend = list(orientation = "h"))
```

```{r}
#| label: bands-nat-method
#| fig-height: 4

d <- scores |>
  filter(scale_id == "natural", !is.na(Method)) |>
  band_share(Method, "natural")
p <- band_plot(d, Method, Method, "Share of forecasts (WIS per 100k band)")
ggplotly(p, tooltip = "text") |> layout(legend = list(orientation = "h"))
```
:::

Model order follows the lookup table (1 = best overall mean log-scale WIS, listed top).

## WIS over time {#view3}

Mean WIS per model structure across the two-year period.

::: {.panel-tabset}
### Log scale

```{r}
#| label: v3-log

d <- scores |>
  filter(scale_id == "log", !is.na(Method)) |>
  group_by(epi_target, Method, target_end_date) |>
  summarise(wis = mean(wis), .groups = "drop")
p <- ggplot(d, aes(target_end_date, wis, colour = Method,
                   text = paste0(Method, "<br>", target_end_date,
                                 "<br>WIS ", round(wis, 2)))) +
  geom_line(aes(group = Method), linewidth = 0.5) +
  facet_wrap(~epi_target, scales = "free_y", ncol = 1) +
  scale_colour_brewer(type = "qual", palette = 2, name = NULL) +
  labs(x = NULL, y = "Mean WIS (log scale)") +
  theme_desc
ggplotly(p, tooltip = "text") |> layout(legend = list(orientation = "h"))
```

### Natural scale

```{r}
#| label: v3-nat

d <- scores |>
  filter(scale_id == "natural", !is.na(Method)) |>
  group_by(epi_target, Method, target_end_date) |>
  summarise(wis = mean(wis), .groups = "drop")
p <- ggplot(d, aes(target_end_date, wis, colour = Method,
                   text = paste0(Method, "<br>", target_end_date,
                                 "<br>WIS ", round(wis, 2)))) +
  geom_line(aes(group = Method), linewidth = 0.5) +
  facet_wrap(~epi_target, scales = "free_y", ncol = 1) +
  scale_colour_brewer(type = "qual", palette = 2, name = NULL) +
  labs(x = NULL, y = "Mean WIS (natural scale)") +
  theme_desc
ggplotly(p, tooltip = "text") |> layout(legend = list(orientation = "h"))
```
:::

## Heatmap over time {#view4}

Colour shows the accuracy band that the model's mean WIS for that week falls in.
Bands here are applied to a mean across countries, not to individual forecasts, so a single extreme forecast can pull a cell into a worse band.

::: {.panel-tabset}
### Log scale

```{r}
#| label: v4-log
#| fig-height: 10

d <- scores |>
  filter(scale_id == "log") |>
  group_by(epi_target, Method, Model, target_end_date) |>
  summarise(wis = mean(wis), .groups = "drop")
ord <- d |> group_by(Method, Model) |>
  summarise(m = mean(wis)) |>
  arrange(desc(m)) |>
  pull(Model)
d <- d |> mutate(lab = factor(id_lab(Model), levels = id_lab(ord)),
                 band = wis_band(wis, "log"))
ggplot(d, aes(target_end_date, lab, fill = band)) +
  geom_tile() +
  facet_wrap(~epi_target) +
  scale_fill_manual(values = band_pal, drop = FALSE,
                    name = "Mean WIS (approx. relative error)") +
  labs(x = NULL, y = "Model") +
  theme_desc +
  theme(axis.text.y = element_text(size = 7))
```

### Natural scale

```{r}
#| label: v4-nat
#| fig-height: 10

d <- scores |>
  filter(scale_id == "natural") |>
  group_by(epi_target, Model, target_end_date) |>
  summarise(wis = mean(wis), .groups = "drop")
ord <- d |> group_by(Model) |> summarise(m = mean(wis)) |> arrange(desc(m)) |> pull(Model)
d <- d |> mutate(lab = factor(id_lab(Model), levels = id_lab(ord)),
                 band = wis_band(wis, "natural"))
ggplot(d, aes(target_end_date, lab, fill = band)) +
  geom_tile() +
  facet_wrap(~epi_target) +
  scale_fill_manual(values = band_pal, drop = FALSE,
                    name = "Mean WIS per 100k") +
  labs(x = NULL, y = "Model No.") +
  theme_desc +
  theme(axis.text.y = element_text(size = 7))
```
:::

## Flagged forecasts

The following death forecasts had natural-scale WIS > 100 per 100k.
Observed deaths per 100k peak around 32.
These flagged forecasts could arise from implementation error, e.g. forecasting cumulative rather than incident weekly values.

```{r}
#| label: broken-table

scores |>
  filter(broken, scale_id == "natural") |>
  count(Model, epi_target, name = "n_forecasts") |>
  arrange(desc(n_forecasts)) |>
  knitr::kable()
```

Forecast scores for deaths at the 1 week horizon per country and week for each flagged model and the ★ ensemble reference, on the natural scale.

```{r}
#| label: drill-setup

drill_models <- unique(c(broken_models, ensemble_model))

# WIS per model x country x week, deaths, natural scale, 1-week horizon (already
# enforced globally on `scores`). Country ordered by that model's own worst week.
drill_dat <- scores |>
  filter(scale_id == "natural", epi_target == "Deaths",
         Model %in% drill_models)

# Static, like the heatmaps above: one plotly trace per tile is not worth the
# render cost when the band already carries the reading.
make_heat <- function(m) {
  d <- drill_dat |> filter(Model == m)
  d <- d |> mutate(Location = factor(as.character(Location)),
                   band = wis_band(wis, "natural"))
  ggplot(d, aes(target_end_date, Location, fill = band)) +
    geom_tile() +
    scale_fill_manual(values = band_pal, drop = FALSE,
                      name = "WIS per 100k") +
    labs(x = NULL, y = NULL) +
    theme_desc +
    theme(axis.text = element_text(size = 7))
}
```

::: {.panel-tabset}
### ILM-EKF ⚠

```{r}
#| label: drill-ilm
#| fig-height: 7

make_heat("ILM-EKF")
```

### epiforecasts-EpiNow2 ⚠

```{r}
#| label: drill-epinow
#| fig-height: 7

make_heat("epiforecasts-EpiNow2")
```

### USC-SIkJalpha ⚠

```{r}
#| label: drill-usc
#| fig-height: 7

make_heat("USC-SIkJalpha")
```

### IEM_Health-CovidProject ⚠

```{r}
#| label: drill-iem
#| fig-height: 7

make_heat("IEM_Health-CovidProject")
```

### EuroCOVIDhub-ensemble ★

```{r}
#| label: drill-ens
#| fig-height: 7

make_heat("EuroCOVIDhub-ensemble")
```
:::

# Observed values: absolute vs per 100k

Underlying observed counts, and the `Incidence` column carried into `m.data`, should be identical for every country-week that survives into the model.
Overlaid lines that coincide confirm no distortion in the pipeline.
Per-100k is `count / population * 100000`.

```{r}
#| label: obs-setup

pop <- read_csv(here("data", "populations.csv")) |> rename(pop = population)

# --- Underlying data: observed-{case,death}.csv ------------------------------
underlying <- c(Cases = "case", Deaths = "death") |>
  purrr::imap(~ read_csv(here("data", paste0("observed-", .x, ".csv"))) |>
                mutate(epi_target = .y)) |>
  bind_rows() |>
  left_join(pop, by = "location") |>
  transmute(
    source = "underlying (observed-*.csv)",
    epi_target, location, target_end_date,
    absolute = observed,
    per100k = observed / pop * 1e5
  )

# --- m.data: the values actually fed to the model ----------------------------
# Incidence = observed count; m.data carries pop, so per-100k is reconstructable.
# Dedupe: Incidence is one value per Location/date/target (repeated across models
# and horizons), so this reads off `scores_all` rather than re-running
# process_data().
mdat <- scores_all |>
  filter(scale_id == "natural") |>
  distinct(epi_target, Location, target_end_date, Incidence, pop) |>
  transmute(
    source = "m.data (Incidence)",
    epi_target, location = as.character(Location), target_end_date,
    absolute = Incidence,
    per100k = Incidence / pop * 1e5
  )

both <- bind_rows(underlying, mdat) |>
  pivot_longer(c(absolute, per100k), names_to = "metric", values_to = "value")
```

## Time series by country

```{r}
#| label: obs-plots
#| results: "asis"
#| fig-width: 11
#| fig-height: 9

plot_one <- function(tgt, met) {
  d <- both |> filter(epi_target == tgt, metric == met)
  ggplot(d, aes(target_end_date, value, colour = source)) +
    geom_line(alpha = 0.8, linewidth = 0.3) +
    facet_wrap(~location, scales = "free_y", ncol = 4) +
    scale_colour_manual(values = c("underlying (observed-*.csv)" = "black",
                                   "m.data (Incidence)" = "red")) +
    labs(title = paste(tgt, "—", met), x = NULL, y = met, colour = NULL) +
    theme_minimal(base_size = 8) +
    theme(legend.position = "top",
          axis.text.x = element_text(angle = 45, hjust = 1))
}

for (tgt in c("Cases", "Deaths")) {
  for (met in c("absolute", "per100k")) {
    cat("\n\n### ", tgt, " — ", met, "\n\n", sep = "")
    print(plot_one(tgt, met))
  }
}
```

## Peak per-100k by country

Highest observed value per country.
Deaths per 100k in the low tens, cases in the low thousands --- both far below the WIS extremes seen on the natural scale.

```{r}
#| label: obs-table

underlying |>
  group_by(epi_target, location) |>
  summarise(max_absolute = max(absolute, na.rm = TRUE),
            max_per100k = round(max(per100k, na.rm = TRUE), 2),
            .groups = "drop") |>
  arrange(epi_target, desc(max_per100k)) |>
  knitr::kable()
```

## Why the natural-scale plots reach values near 100,000

Those are WIS values, not observed deaths.
WIS is computed on the per-100k scale (`R/process-score.R` scales `observed`/`prediction` by `/pop * 1e5`).
A WIS near 100,000 means a model's predicted interval was \~100,000 per-100k away from the observed value --- an implausible forecast, since observed deaths per 100k peak around 30.
The extreme is a broken forecast submission, not a data-scaling error.

```{r}
#| label: wis-vs-obs

scores_all |>
  filter(scale_id == "natural", !grepl("EuroCOVIDhub-", Model)) |>
  group_by(epi_target) |>
  summarise(
    max_observed_per100k = round(max(Incidence / pop * 1e5, na.rm = TRUE), 1),
    max_wis = round(max(wis), 1),
    .groups = "drop"
  ) |>
  knitr::kable()
```
