---
title: "Supplement"
output:
  pdf_document:
    toc: true
---

```{r set-up, include=FALSE}
library(here)
library(dplyr)
library(readr)
library(tidyr)
library(ggplot2)
library(forcats)
library(patchwork)
library(flextable)
theme_set(theme_bw())
knitr::opts_chunk$set(eval = TRUE, echo = FALSE,
                      message = FALSE, warning = FALSE)
```

```{r load-data}
# Load data
source(here("R", "prep-data.R"))
source(here("R", "descriptive.R"))
scores <- prep_data(scoring_scale = "log")
```
 
## Team scores and metadata

```{r metadata-table}
metadata <- read_csv(here("data", "model-classification.csv")) |>
  select(Model = model, Method = classification)
affiliation <- scores |> 
  distinct(Model, country_affiliation) |> 
  rename(AffiliatedLocation = country_affiliation)
targets <- table_targets(scores) |> 
  select(Model, CountryTargets)
model_scores <- scores |>
  group_by(Model) |> 
  table_confint() |> 
  select(Model, Forecasts, median, `Median (IQR)`)
metadata |> 
  left_join(targets |> rename("Country Targets" = CountryTargets)) |> 
  left_join(affiliation |> rename("Affiliated Location" = AffiliatedLocation)) |> 
  left_join(model_scores) |> 
  arrange(median, Method, `Country Targets`, `Affiliated Location`) |> 
  mutate(Description = paste0("[Metadata](https://raw.githubusercontent.com/covid19-forecast-hub-europe/covid19-forecast-hub-europe/main/model-metadata/", Model, ".yml)")) |> 
  select(-median) |> 
  knitr::kable()
```

\newpage

## Model fitting

```{r model-wis}
source(here("R", "model-wis.R"))
```
We fit a generalised additive mixed effects model using the `mgcv` package in `R`. Code is available at: <https://github.com/epiforecasts/model-structure-evaluation>. 

#### 1. Conditional distribution of outcomes

```{r density-plot, fig.height=6, fig.width=5}
plot_density(scores)
```

#### 2. Model formula

`r m.fit$formula`

\newpage

#### 3. Summary of estimates

Reference levels for categorical variables are:
- Mechanistic model method
- Single-country target
- Modelling team affiliation is located in the forecast target country 
- Trend (of observed incidence) is stable
- Horizon is one-week-ahead

```{r gamm-coeffs-all}
as_flextable(m.fit) |> 
  autofit()
```

```{r print-gamm}
knitr::kable(m.anova$pTerms.table, digits = 3, caption = "ANOVA for parametric terms")
```

\newpage

#### 4. Partial effects plots


```{r gamm-plot-all, fig.height=5,fig.width=5, fig.cap="Partial residual plots for all terms"}
draw(m.fit, parametric = TRUE, residuals = TRUE, ci_level = 0.95)
```
\newpage

#### 5. Model diagnostics

```{r gamm-diagnostics, fig.height=5,fig.width=5}
# QQ plot, residuals
appraise(m.fit)
```

  
```{r scores-over-time, fig.cap=plot_iqr_over_time_cap, fig.height=3, fig.width=5}
# plot_iqr_over_time_cap <- "Median (IQR) of the weekly weighted interval score across all forecasts submitted to the European COVID-19 Forecast Hub"
# plot_iqr_over_time(scores)
```
