---
title: "README"
output: github_document
---

### Data processing

The main script for combining and preparing data for analysis is: [`R/process-data.R`](R/process-data.R). The following R scripts were used to process the data stored here:

- [`R/utils-data.R`](R/utils-data.R) : Code to access forecasts, observed data, and population size for each location
  - Outputs:
     - `populations.csv`
     - `observed-case.csv`, `observed-death.csv`
     - `covid19-forecast-hub-europe.parquet`
- [`R/utils-metadata.R`](R/utils-metadata.R) : Functions to access model names, model submissions, and raw metadata from Hub GitHub repos.
  - Output: `model-classification.csv`
- [`R/utils-variants.R`](R/utils-variants.R) : Takes raw data on variant sequencing stored here, cleans and classifies phases of dominant variants
  - Inputs:
    - `variants/eu-ecdc.csv`: 30 EU countries covered by the ECDC, source: https://www.ecdc.europa.eu/en/publications-data/data-covid-19-variants-eueea
    - `variants/uk-ukhsa.csv`: United Kingdom, source: https://ukhsa-dashboard.data.gov.uk/respiratory-viruses/covid-19
    - `variants/ch-wgs.csv` and `variants/ch-hosp.csv`: Switzerland, source: https://covid19.admin.ch/
- [`R/process-score.R`](R/process-score.R) : Code to score forecasts using scoringutils
  - Outputs: 
    - `scores-raw-case.csv`
    - `scores-raw-death.csv`


#### Model metadata

Code used to access and store metadata:

```{r, eval=FALSE}
library(here)
library(purrr)
library(lubridate)
source(here("R", "utils-metadata.R"))

# set repo paths
repos <- c(
  "euro" = "covid19-forecast-hub-europe/covid19-forecast-hub-europe",
  "us" = "reichlab/covid19-forecast-hub"
)

# date range for submissions
min_date <- "2021-03-07"
max_date <- as.Date("2023-03-10") - weeks(4)

# get euro metadata
eu_submissions <- get_model_submissions(
  repo = repos[["euro"]],
  min_date = min_date,
  max_date = max_date
)
eu_metadata_raw <- get_metadata_raw(
  repo = repos[["euro"]],
  model_abbr = unique(eu_submissions$model)
)
write_metadata(
  metadata_raw = eu_metadata_raw,
  sheet_name = "euro-metadata-raw", write_local = FALSE
)

```

Code can also be used to access US Hub metadata:

```{r, eval=FALSE}
# get US metadata
us_submissions <- get_model_submissions(
  repo = repos[["us"]],
  min_date = min_date,
  max_date = max_date
)
us_metadata_raw <- get_metadata_raw(
  repo = repos[["us"]],
  model_abbr = unique(us_submissions$model)
)

us_metadata_raw <- us_metadata_raw |>
  mutate(euro_overlap = model_abbr %in% unique(eu_submissions$model))

write_metadata(
  metadata_raw = us_metadata_raw,
  sheet_name = "us-metadata-raw", write_local = FALSE
)
```
