## Model structure classification

Human raters classified models according to methodological structures.

```{r}
#| label: tbl-supp-classification
#| tbl-cap: >-
#|   Classification of models by number of raters in total and agreement on model structure.

model_classifier <- classify_models(return_majority = FALSE) |> 
  distinct(model, classification, raters, votes, agreement) |> 
  group_by(model) |>
  mutate(final = classification[which.max(votes)]) |> 
  pivot_wider(names_from = classification, values_from = votes, values_fill = 0) |> 
  filter(model %in% unique(scores$Model)) |>
  select(Model = model, `Final classification` = final,
         Agreement = agreement, `Total raters` = raters, everything()) |>
  arrange(Agreement, Model)

model_classifier |> 
    kable()
```

## Epidemic trend identification

We categorised each week as "Stable", "Decreasing", or "Increasing", based on the difference over a three-week moving average of incidence (with a change of +/-5% as "Stable").

```{r}
#| label: fig-supp-trends
#| fig-cap: >-
#|   Weekly incidence of reported cases in each country, coloured by the epidemic
#|   trend assigned to that week ("Stable", "Increasing" or "Decreasing"), from the
#|   change in a three-week moving average.
#| fig-height: 8
#| fig-width: 10

trends_plot <- function(scores) {
    trends <- scores |>
        select(Location, target_end_date, Incidence, Trend) |>
        distinct()
    p <- ggplot(trends, aes(x = target_end_date, y = Incidence)) +
        geom_point(mapping = aes(colour = Trend), size = 1) +
        geom_line() +
        scale_colour_brewer(palette = "Set2", na.value = "grey") +
        theme(legend.position = "bottom") +
        facet_wrap(~Location, scales = "free_y") +
        theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
        xlab("")
    return(p)
}

scores |>
  filter(epi_target == "Cases") |>
  trends_plot() +
  ggtitle("Cases")
```

```{r}
#| label: fig-supp-death-trends
#| fig-cap: >-
#|   Weekly incidence of reported deaths in each country, coloured by the epidemic
#|   trend assigned to that week, derived as for cases.
#| fig-height: 8
#| fig-width: 10

scores |>
  filter(epi_target == "Deaths") |>
  trends_plot() +
  ggtitle("Deaths")
```

## Variant phase identification

Genomic surveillance data were obtained from three sources: ECDC (covering 30 European countries), UKHSA (Great Britain), and the Swiss Federal Office of Public Health (Switzerland).
Variant lineages were mapped to six named phases in expected chronological order: Alpha, Delta, Omicron-BA.1, Omicron-BA.2, Omicron-BA.4/5, and Omicron-BQ/XBB.
For each country, we identified the first week in which each named variant exceeded 50% of sequenced samples.
We enforced chronological ordering by removing any out-of-sequence phases, then expanded phase assignments to all weeks by filling forward and backward from observed transition dates.
This per-location approach accounts for the fact that variant dominance dates differed substantially across European countries.
Where genomic surveillance data were too sparse to identify a transition (Hungary), we supplemented with epidemiological reports to set the Alpha-to-Delta transition date.

```{r}
#| label: fig-supp-variant-phases
#| fig-cap: "Variant phases identified by dominant variant in each location and week"

# see: R/utils-variants.R
variant_phases <- classify_variant_phases()
variant_phases |>
    ggplot(aes(x = target_end_date, y = location, fill = VariantPhase)) +
    geom_tile() +
    scale_fill_brewer(type = "qual", palette = 2) +
    labs(x = NULL, y = "Location", fill = "Variant phase") +
    theme_minimal() +
    theme(legend.position = "bottom")
```
