Effects are random effects for each level of a covariate, relative to a forecast at the average level of that covariate with all other terms held fixed.
Negative raw effects, or ratios below 1, indicate better-than-average performance.

## Partial effects across covariates

Partial effect on forecast performance of key covariates included in the fully adjusted model for additional covariates.
Partial effects of individual model structure and individual model are presented in the main text.

(A) Geographic

```{r}
#| label: supplement-plot-spatial-effects
#| fig-cap: >-
#|   Performance varied more between countries than between single- and
#|   multi-country models. Partial effects (95% CI) for geographic scope and
#|   country, as a ratio of the expected LWIS relative to the average level.
# TODO. in supplement, panel target features into a single figure (A: epi outcome, B: epidemic trend, C: variant phase, D: country), next to it panel forecaster features (A: single/multi country, B: model structure); show adjusted effect only; use a sensible colour scale to differentiate groups. 
plot_effects(results$effects, variables = c("CountryTargets", "Location"))
```

(B) Epidemic state

```{r}
#| label: supplement-plot-temporal-effects
#| fig-cap: >-
#|   Epidemic trend and variant phase both shifted performance, with increasing
#|   trends and the Omicron BA.1 phase hardest to predict. Partial effects
#|   (95% CI) as a ratio of the expected LWIS relative to the average level.

plot_effects(results$effects, variables = c("Trend", "VariantPhase"))
```

## Partial effects on the raw scale

The main-text figures and tables report exponentiated effects (multiplicative ratios of the expected LWIS relative to the average level of each factor).
The table below gives the underlying raw partial effects from the fully adjusted model.

```{r}
#| label: supplement-raw-coefficients

results$effects |>
  filter(model == "Adjusted") |>
  transmute(
    Variable = group_var,
    Group = group,
    `Partial effect (95% CI), log scale` = paste0(
      round(value, 3),
      " (", round(lower_2.5, 3), ", ", round(upper_97.5, 3), ")"
    )
  ) |>
  arrange(Variable, Group) |>
  kable(
    caption = paste0(
      "Raw partial effects on the log scale of the linear predictor, from ",
      "the fully adjusted generalised additive mixed model. Negative values indicate ",
      "better-than-average performance. 95% CI = 95% confidence interval."
    ),
    align = c("l", "l", "r")
  ) |>
  collapse_rows(columns = 1, valign = "top") |>
  kable_styling(full_width = FALSE)
```

The model structure term is fitted as a single grouping factor crossing structure with epidemiological outcome, so its raw partial effects are one per structure-by-outcome cell.
The main text reports these exponentiated, and reports a pooled effect per structure derived as a contrast averaging a structure's two cells.

```{r}
#| label: supplement-table-method-target

print_table_method_target(results$method_by_target)
```

```{r}
#| label: supplement-method-target-raw

results$method_by_target |>
  transmute(
    `Model structure` = Method,
    Outcome = Epi_target,
    `Partial effect (95% CI), log scale` = paste0(
      round(value, 3),
      " (", round(lower_2.5, 3), ", ", round(upper_97.5, 3), ")"
    )
  ) |>
  arrange(`Model structure`, Outcome) |>
  kable(
    caption = paste0(
      "Raw partial effects of model structure by epidemiological outcome, ",
      "on the log scale, from the fully adjusted generalised additive mixed ",
      "model. Negative values indicate better-than-average performance. ",
      "95% CI = 95% confidence interval."
    ),
    align = c("l", "l", "r")
  ) |>
  collapse_rows(columns = 1, valign = "top") |>
  kable_styling(full_width = FALSE)
```

## Model ranking before and after adjustment

Individual models can be ranked by their partial effect on the score, either from a univariate model containing only individual model identity, or from the fully adjusted model.
The first ranks models by observed performance, and so by a combination of the method used and the difficulty of the targets each model chose to forecast.
The second ranks them by performance after the target-related covariates are held fixed.
The comparison is reported in the main text; the table below gives the models that move furthest between the two.

```{r}
#| label: supplement-ranks

ranks <- rank_models(results$effects)
rank_summary <- summarise_ranks(ranks)

ranks |>
  mutate(`Rank change` = rank_change) |>
  arrange(desc(abs(`Rank change`))) |>
  head(10) |>
  transmute(
    Model = label,
    `Unadjusted rank` = rank_unadjusted,
    `Adjusted rank` = rank_adjusted,
    `Rank change` = `Rank change`
  ) |>
  kable(
    caption = paste0(
      "The ten models whose rank changes most under adjustment, of ",
      rank_summary$n, " models. A positive change means a model ranks better ",
      "once the difficulty of its targets is accounted for. Models are ",
      "anonymised and numbered within model structure."
    ),
    align = c("l", "r", "r", "r")
  ) |>
  kable_styling(full_width = FALSE)
```
